---
title: "Dataset description"
bibliography: references.bib
---

# Dataset building

## 1. Environmental data

### 1.1. Deforestation {#sec-deforestation}

#### 1.1.1. Understand the data

The deforestation data used in this study comes from the **MapBiomas Project**, a Brazilian initiative that provides annual land use and land cover data for the entire national territory. The data is derived from Landsat satellite imagery with 30-meter resolution, processed pixel by pixel to produce consistent information across all Brazilian biomes.

I use the *Deforestation and Secondary Vegetation Database*, *version 9*, which covers the period from 1985 to 2023 (last updated in August 2024, available at <https://brasil.mapbiomas.org/estatisticas/>).

The database provides the area (in hectares) of primary and secondary total vegetation and vegetation supression per year for Brazilian municipalities. Land use is categorized into six classes and organized across four levels of detail. For this research, I filter the data based on four land use classes:

i.  *Primary Vegetation:* Indicates absence of deforestation event, i.e., permanence since the base year in one or more classes of natural vegetation

ii. *Secondary Vegetation:* Indicates trajectory with the presence of a recovery for deforestation event in previous years.

iii. *Primary Vegetation Suppression:* Indicates a deforestation event occurring in year $t$, in a pixel previously classified as *Primary Vegetation*. In the following year $(t+1)$, the pixel is reclassified as *Anthropic Use*.

iv. *Secondary Vegetation Suppression:* Indicates a deforestation event in year $t$, in a pixel previously classified as *Secondary Vegetation*, which is reclassified as *Anthropic Use* in year $t+1$.

Regarding the levels of land use detail (e.g., forest formation, savana formation, floodable forest, and grassland formationt), I use only *Level 1*, the most general level, as my objective is to analyze overall patterns of deforestation.

#### 1.1.2. Import and inspect the data

I download the data of *Deforestation and Secondary Vegetation* from the *Statistics* section of the Mapbiomas website. I inspect the data to ensure that the import function correctly interpret the decimal separator and recognize numbers as numeric or double values.

The raw dataset contains 51 columns, including *country, state, municipality name, municipal geocode,* and the *transition category* (transition_name). It also includes four columns representing *land-use* categories and 38 columns corresponding to individual *years*, each containing the value of deforestation for that municipality and land-use category.

With the initial inspection, I also discover that the values for *Primary Vegetation Suppression* for the years of 1985 and 1986 are all zero, which justifies the exclusion of those years from the dataset.

```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(readr)
library(writexl)
library(ggplot2)

data_deforestation_raw <- read_xlsx("_data/raw-environ-mapbiomas-deforestation-municipalities.xlsx", sheet = 2)

data_deforestation_raw %>%
  filter(transition_name == "Supressão Veg. Primária") %>%
  summarise(all_zero = all(`1985` == 0)) 

# I repeated this check for the years 1986, 1987, and 2023

data_deforestation_raw %>%
  filter(transition_name == "Supressão Veg. Primária") %>%
  summarise(non_zero_count_1985 = sum(`1985` != 0), 
            non_zero_count_1986 = sum(`1986` != 0),
            non_zero_count_1987 = sum(`1987` != 0),
            non_zero_count_2023 = sum(`2023` != 0))
```

#### 1.1.3. Clean and transform the data

I select relevant columns: *state*, *municipality*, *geocode*, *transition name*, and *area* in hectares for each year from 1987 to 2023, filter for Amazon states and the specific transitions that I am interested in (primary and secondary vegetation and their respective suppression), recode transition names, and reshape the data to a tidy format. I convert the *year* values to integers to enable numeric operations, and the *geocode* values to characters to prevent them from being treated as numbers. Finally, I check if all the *geocode* are correctly registered with seven digits.

```{r}
data_deforestation_clean <- data_deforestation_raw %>%
  select(
    state, municipality, geocode, transition_name, `1987`:`2023`
  ) %>% 
  filter(
    state %in% c("Acre", "Amapá", "Amazonas", "Maranhão", "Mato Grosso", "Pará", "Rondônia", "Roraima", "Tocantins"),
    transition_name %in%  c("Supressão Veg. Primária", "Supressão Veg. Secundária", "Veg. Primária", "Veg. Secundária")
  ) %>%
  mutate(
    transition_name = recode(
      transition_name,
      "Supressão Veg. Primária" = "loss_of_primary_vegetation",
      "Supressão Veg. Secundária" = "loss_of_secondary_vegetation",
      "Veg. Primária" = "primary_vegetation",
      "Veg. Secundária" = "secondary_vegetation"),
    geocode = as.character(geocode)
  ) %>%
  pivot_longer(
    cols = `1987`:`2023`,
    names_to = "year",
    values_to = "area_ha",
    names_transform = list(year = as.integer)
  )

data_deforestation_clean %>%
  filter(nchar(geocode) == 7) %>% 
  count()
```

As my interest is in the total forest loss (the sum of primary and secondary vegetation loss), I create the new column *vegetation_category* where I categorize the vegetation variables only as *forest* (sum of primary and secondary vegetation) and *deforestation* (sum of loss of primary and secondary vegetation). I summarize the total area in hectares for each vegetation category by municipality, and reshape the data to a wide format.

```{r}
data_deforestation_tidy <- data_deforestation_clean %>%
  mutate(
    vegetation_category = case_when(
      transition_name %in% c("primary_vegetation", "secondary_vegetation") ~ "forest",
      transition_name %in% c("loss_of_primary_vegetation", "loss_of_secondary_vegetation") ~ "deforestation",
      TRUE ~ NA_character_
    )
  ) %>%  
  filter(!is.na(vegetation_category)) %>%
  group_by(
    state, municipality, geocode, year, vegetation_category
  ) %>%
  summarise(
    total_area_ha = sum(area_ha, na.rm = TRUE),
    .groups = 'drop'
  ) %>% 
  pivot_wider(
    id_cols = c(state, municipality, geocode, year),
    names_from = vegetation_category,
    values_from = total_area_ha,
    values_fill = 0,
    names_prefix = "area_"
  )
```

#### 1.1.4. Check and validate the data

I assess the dataset for missingness and find none. I also analyze the dataset for outliers using summary statistics and graphs. The data is skewed to the right, as expected, since the dataset currently includes municipalities across all Amazon states, including many with very low deforestation rates. I checked for duplicates and found none, indicating that the data is clean and that each record is unique.

Finally, I cross-validate the dataset values against the online dashboard data available on the Mapbiomas website (`https://brasil.mapbiomas.org/` hit the "Maps and Data" button, then "Access the Platform", and "Mapbiomas Land Cover and Use platform"; select "Deforestment" in the left menu). The intention here is to evaluate whether the downloaded data is consistent with what is released online by Mapbiomas and whether I have correctly managed the transformations. I randomly selected municipalities to compare their values with those shown on the MapBiomas platform. As my dataset summarize the values for primary and secondary vegetation, I manually calculated the same sums using the values available on the platform. The validation confirms that the values in my dataset align with those available on the MapBiomas platform.

#### 1.1.5. Correct and standardize units

I perform two standardization transformations to smooth the cross-sectional variation in deforestation arising from heterogeneity in municipality size.

First, I calculate the deforestation rate as the percentage of deforested area relative to the total forest area for each municipality-year observation. This enables a proportional comparison of deforestation intensity across units with varying forest sizes.

The measure can be expressed mathematically as:

$$
\text{DefRate}_{it} = \frac{\text{DefArea}_{it}}{\text{ForestArea}_{it}} \times 100
$$

where $DefArea_{it}$ is the deforested area in municipality $i$ at time $t$, and $ForestArea_{it}$ is the total forest area of municipality $i$ in the same year.

Second, I compute the z-scores for deforestation areas within each municipality, standardizing values by subtracting the municipal mean and dividing by the municipal standard deviation. Z-scores indicate how far each value deviates from the municipality’s average in standard deviation units. Positive values represent above-average deforestation, while negative values indicate below-average levels.

The measure can be expressed mathematically as:

$$
Z_{it} = \frac{D_{it} - \mu_i}{\sigma_i}
$$

where $D_{it}$ is the deforestation area in municipality $i$ at time $t$, $\mu_i$ is the mean deforestation area of municipality $i$, and $\sigma_i$ is the corresponding standard deviation.

```{r}
data_deforestation_standardized <- data_deforestation_tidy %>%
  mutate(
    rate_deforestation = if_else(
      area_forest > 0,
      (area_deforestation / area_forest) * 100,
      NA_real_
    )
  ) %>%
  group_by(geocode) %>%
  mutate(
    mean_deforestation = mean(area_deforestation, na.rm = TRUE),
    sd_deforestation = sd(area_deforestation, na.rm = TRUE),
    z_deforestation = (area_deforestation - mean_deforestation) / sd_deforestation
  ) %>%
  ungroup()
```

## 1.2. Biome

#### 1.2.1. Understand the data

I use the Predominant Biome by Municipality for Statistical Purposes list (**Bioma Predominante por Município para Fins Estatísticos**) to determine which municipalities I should include in the dataset. The data on biome by municipalities was released by the Brazilian Institute of Geography and Statistics (IBGE) in June 2024 (avaiable at <https://www.ibge.gov.br/geociencias/cartas-e-mapas/informacoes-ambientais/15842-biomas.html?=&t=downloads>).

This list identifies the predominant biome in each Brazilian municipality, providing a standardized criterion for aggregating municipal data by biome. Municipalities located in more than one biome are classified according to the biome that covers the largest proportion of their territorial area.

In this study, I use the Amazon biome as a filter to select the municipalities to be analyzed. According to the IBGE classification, the Amazon is the predominant biome in 503 Brazilian municipalities.

#### 1.2.1. Import and inspect the data

I import and inspect the data to check the number of municipalities registered with "Amazônia" as the predominant biome, and correctly find 503 entries.

```{r}
data_biome_raw <- read_csv2("_data/raw-environ-ibge-biome.csv", locale = locale(encoding = "UTF-8"))

unique(data_biome_raw$`Bioma predominante`)

data_biome_raw %>%
  count(`Bioma predominante`) %>%
  arrange(desc(n))
```

#### 1.2.3. Clean and tidy the data

I rename the columns and selected the ones of interest.

```{r}
data_biome_tidy <- data_biome_raw %>%
  rename (
    geocode = `Geocódigo`,
    municipality = `Nome do município`,
    state = `Sigla da UF`,
    biome = `Bioma predominante`
  ) %>%
  mutate(geocode = as.character(geocode))

data_biome_tidy %>%
  filter(biome == "Amazônia") %>%
  distinct(geocode) %>%
  count()
```

#### 1.2.4. Join Deforestation and Biome datasets

I join the biome data with the deforestation dataset using the geocode to match them. I use the *municipality* column to check if is there mismatch between the municipality information in both datasets, find only one related to a typo in the spelling of a municipality name and keep the column with the correct name. I tidy and clean the final dataset and filter by the *biome* of my interest, the Amazon biome.

```{r}
data_deforestation_with_biome <- data_deforestation_standardized %>%
  left_join(data_biome_tidy, by = "geocode")

mismatch_check <- data_deforestation_with_biome %>%
  filter(municipality.x != municipality.y)

view(mismatch_check)

data_deforestation_final <- data_deforestation_with_biome %>%
  rename(
    municipality = municipality.x,
    state = state.x,
    state_abbreviation = state.y,
  ) %>%
  select(-municipality.y) %>% 
  select(
    geocode, 
    municipality, 
    state,
    state_abbreviation, 
    biome, 
    year, 
    area_forest, 
    area_deforestation, 
    rate_deforestation, 
    z_deforestation,
    mean_deforestation,
    sd_deforestation
  ) %>%
  filter(biome == "Amazônia")
```

I count the number of unique geocodes and municipalities, which was supposed to be 503, as the IBGE classification. I correctly find 503 unique geocodes, but only 499 municipalites. My inspection pointed out that this is because some municipalities in different states can share the same name.

```{r}
data_deforestation_final %>%
  distinct(geocode) %>%
  count()

data_deforestation_final %>%
  distinct(municipality) %>%
  count()
```

The municipalities with the same name are:

| \#  | geocode | municipality      | state |
|-----|---------|-------------------|-------|
| 1   | 2100873 | Araguanã          | MA    |
| 2   | 1702158 | Araguanã          | TO    |
| 3   | 1505551 | Pau D'Arco        | PA    |
| 4   | 1716307 | Pau D'Arco        | TO    |
| 5   | 2109239 | Presidente Médici | MA    |
| 6   | 1100254 | Presidente Médici | RO    |
| 7   | 1200401 | Rio Branco        | AC    |
| 8   | 5107206 | Rio Branco        | MT    |


### 1.3. PPCDAm

#### 1.3.1. Understand the data

The data used to identify municipalities subject to intensified environmental enforcement under the PPCDAm are derived from a set of official normative acts issued by the Brazilian federal government between 2007 and 2013. These include the [Decreto Nº 6.321](https://www.planalto.gov.br/ccivil_03/_ato2007-2010/2007/decreto/d6321.htm) and subsequent ministerial ordinances (Portarias) published by the Ministry of the Environment (MMA), which operationalized the policy by defining and updating the list of *priority municipalities*.

The *PPCDAm* (*Plano de Ação para Prevenção e Controle do Desmatamento na Amazônia Legal*) introduced a targeted enforcement strategy based on the identification of municipalities with critical deforestation dynamics. The initial legal framework established by the decree defined criteria for inclusion in the list, which typically combined:

- High absolute levels of deforestation;
- Rapid recent increases in deforestation;
- Persistent deforestation over time.

Subsequent ordinances periodically updated the list by adding new municipalities, maintaining those that continued to meet the criteria, and, in some cases, removing municipalities that demonstrated sustained reductions in deforestation and compliance with environmental regulations. All the ordinances could be found here (XXXXX)

Inclusion in the list triggered a set of stricter command-and-control measures, including:

- Intensified environmental monitoring and inspections;
- Restrictions on access to rural credit;
- Embargoes on properties associated with illegal deforestation;
- Increased coordination among federal enforcement agencies.

### 1.3.2 Data construction and coding

The information contained in the decrees and ordinances was manually compiled and harmonized into a municipality-year panel dataset. Each legal document was reviewed to extract the list of municipalities classified as priority in the corresponding year.

Given that the ordinances were issued at different points in time and sometimes updated the list within the same year, the dataset was standardized to an annual frequency. For each municipality $i$ and year $t$, a binary indicator was constructed:

$$
PPCDAm_{it} =
\begin{cases}
1 & \text{if municipality } i \text{ is included in the priority list in year } t \\
0 & \text{otherwise}
\end{cases}
$$

This coding implies the following operational decisions:

- **Treatment group (1):** municipalities listed as priority in a given year;
- **Control group (0):** municipalities not listed in that year;

### 1.3.3. Clean and transform the data

The original spreadsheet was organized horizontally, then I reshaped it into long format for the purposes of panel data analysis. 

After reshaping, the resulting dataset contains at least four core variables:

- `municipality_name`: municipality name;
- `ibge_code`: official IBGE municipality code;
- `year`: reference year;
- `ppcdam_list`: binary indicator of PPCDAm priority-list status in that year.


```{r}
library(dplyr)
library(tidyr)
library(readr)

ppcdam_long <- ppcdam_wide %>%
  pivot_longer(
    cols = starts_with("ppcdam_"),
    names_to = "year",
    names_prefix = "ppcdam_",
    values_to = "ppcdam_list"
  ) %>%
  mutate(
    year = as.integer(year),
    ibge_code = as.numeric(ibge_code),
    ppcdam_list = as.integer(ppcdam_list)
  ) %>%
  arrange(ibge_code, year)
```


## 2. Social data

### 2.1. Employment

#### 2.1.1. Understand the data

The employment data used in this study comes from the **Relação Anual de Informações Sociais (RAIS)**, an administrative dataset maintained by the Brazilian Ministry of Labor and Employment. RAIS compiles annual information reported by all formal-sector employers in Brazil, covering private firms, public agencies, and third-sector organizations. The dataset includes detailed records of each employment bond, such as the worker’s occupation, income, education level, sex, age, and location of employment. Because reporting to RAIS is mandatory, it provides near-complete coverage of formal employment relationships in the country.

The dataset spans 1985–2024 and is divided into two main components:

*Estabelecimentos*, containing establishment-level information, and *Vínculos*, containing individual employment-bond microdata.

This study utilizes the microdados *Vínculos*, where each row corresponds to a unique employment bond in a given year. Variables describe the characteristics of the worker (e.g., education, sex, age), the job (e.g., wages, occupation), and the establishment. The data are anonymized.

Due to its depth and temporal coverage, RAIS enables consistent measurement of formal employment trends at fine geographic scales, including municipalities within the Brazilian Amazon.

#### 2.1.2. Import and inspect the data

The RAIS microdata contain more than 2 billion rows, 350 GB, making local processing difficult. Therefore, I conduct all data manipulation using **Google BigQuery**, a distributed cloud data warehouse that enables scalable SQL processing and avoids local storage constraints.

The data were accessed through **Base dos Dados (BD)**, a non-profit open-data infrastructure that republishes harmonized versions of major Brazilian administrative datasets @dahis2022. BD performs several preprocessing procedures to guarantee temporal consistency and standardization across years, including:

-   Converting municipality identifiers to the official 7-digit IBGE geocode;

-   Standardizing state identifiers to UF acronyms;

-   Harmonizing classification systems (CNAE, CBO, establishment types) across regulatory changes;

-   Replacing invalid or placeholder values (e.g., “9999”, “000”) with nulls.

The dataset I use corresponds to the BigQuery table *basedosdados.br_me_rais.microdados_vinculos*, with the latest update as of 2025-08-26 (accessed November 14, 2025). The dataset could be acessed through the BD website (<https://basedosdados.org/dataset/3e7c4d58-96ba-448e-b053-d385a829ef00?table=dabe5ea8-3bb5-4a3e-9d5a-3c7003cd4a60>) and directly in the Google BigQuerry storage (<https://console.cloud.google.com/bigquery?ws=!1m5!1m4!4m3!1sbasedosdados!2sbr_me_rais!3smicrodados_vinculos>). The BD cleaning scripts are publicly available in the associated GitHub repository (<https://github.com/rdahis/clean_RAIS>).

Before proceeding to filtering and transformation, I performed an initial inspection of the dataset to check schema consistency, variable types, missingness patterns, and key descriptive distributions (e.g., number of vínculos by year, UF, and sector). It was possible to build some visualizations and graphs directly in Google BigQuery, using the Looker Studio. This confirmed the integrity of the BD harmonized structure.

#### 2.1.3. Filter and transform the data

I construct annual municipal indicators of formal employment for the Legal Amazon region from the RAIS *Vínculos* microdata. All steps described below were implemented through SQL queries in BigQuery, following a transparent, reproducible workflow. For clarity, the data transformation process is organized into six components:

##### 2.1.3.1. Spatial filtering

Municipality (*id_municipio* column in the original dataset): the variable *id_municipio*, already standardized to the 7-digit IBGE format by BD, was used as the primary geographic identifier.

State selection (*sigla_uf* column in the original dataset): only municipalities located in the Legal Amazon were included (AC, AM, AP, MA, MT, PA, RO, RR, TO).

##### 2.1.3.2. Temporal filtering

To maintain consistency in key variables, only data from 1999 to 2024 (*ano* column in the original dataset) are used. Earlier years were excluded because they present limitations in occupational coding (CBO).

##### 2.1.3.3. Employment-status filtering

The variable identifies whether the work bond was active on December 31 of each year. Only bonds with *vinculo_ativo_3112* $= 1$ (meaning ‘active’) were retained to avoid duplicates and produce consistency.

##### 2.1.3.4. Occupational categories

RAIS uses the Brazilian Classification of Occupations/Classificação Brasileira de Ocupações (CBO) to code worker occupations. Over the study period, two versions of CBO were used, corresponding to the classification system used until 2002 (*cbo_1994*) and the one used after (*cbo_2002*). To harmonize them:

i.  When available, the 6-digit *cbo_2002* code was used directly.

ii. When only 5-digit *cbo_1994* was available, the code was converted to a 6-digit format by appending a zero.

iii. The first number of both codes are related to mayor's group of the CBO. The category “Agricultural, Forestry, and Fishing Worker” has the code 6. To simplify analysis and produce a consistent agricultural–non-agricultural divide across the study period, a binary occupational variable was created:

-   Agricultural workers (1): occupations whose unified CBO code begins with “6”
-   Non-agricultural workers (0): all other occupations

##### 2.1.3.5. Deflation and wage groups {#sec-deflation}

The variable *valor_remuneracao_media* provides the worker’s annual average wage. Only positive, non-null wage values were kept. To allow temporal comparability, all wage values were **deflated to 2024** Brazilian reais using the **IPCA** national consumer price index (IBGE).

Formally, the deflator for each year $t$ is defined as:

$$
Deflator(t) = \frac{IPCA(2024)}{IPCA(t)}
$$

The deflator table (constructed separately using a simple spreadsheet tool) was merged with RAIS data. Deflated wages were then calculated as:

$$
Wage^{2024}_{it} = Wage_{it} \times Deflator(t)
$$

Data on the IPCA historic series can be found in the IBGE website: https://www.ibge.gov.br/estatisticas/economicas/precos-e-custos/9256-indice-nacional-de-precos-ao-consumidor-amplo.html?=&t=downloads (accessed November 14, 2025).

The deflator table constructed and merged with RAIS data in BigQuerry can be accessed at the following link: *XXXXXXXXXXXXXXXXXXXX*

Deflated wages were categorized into six groups using the 2024 minimum wage (R\$ 1,412.00) as the reference:

1.  Up to 1 minimum wage

2.  From 1 to 2 minimum wages

3.  From 2 to 3 minimum wages

4.  From 3 to 4 minimum wages

5.  From 4 to 5 minimum wages

6.  More than 5 minimum wages

These categories capture the wage distribution at the base of the income pyramid and facilitate the interpretation of a phenomenon that I want to observe, which is the policy of increasing the minimum wage over the last 25 years.

##### 2.1.3.6. Aggregation

A final SQL query aggregated the processed microdata into municipal-year indicators by:

-   year,
-   municipality (geocode),
-   occupational group (agricultural vs. non-agricultural),
-   wage group, and
-   total number of formal workers in each subgroup.

Two SQL queries were executed in Google BigQuery to conduct all filtering, harmonization, deflation, and aggregation procedures. For transparency and reproducibility, both scripts are documented below.

The **first query** filters the raw data, harmonizes occupation categories, and applies the wage deflation procedure:

```{sql connection=}

SELECT
  r.ano,
  r.sigla_uf,
  r.id_municipio,
  CASE
  WHEN r.cbo_2002 IS NOT NULL THEN CAST(r.cbo_2002 AS STRING)
  WHEN r.cbo_1994 IS NOT NULL THEN CONCAT(CAST(r.cbo_1994 AS STRING), '0')
  ELSE NULL
END AS cbo_unified,
  r.vinculo_ativo_3112,
  r.valor_remuneracao_media,
  d.ipca_deflator,
  SAFE_MULTIPLY(r.valor_remuneracao_media, d.ipca_deflator) AS valor_remuneracao_media_2024
FROM
  `basedosdados.br_me_rais.microdados_vinculos` AS r
LEFT JOIN
  `amz-data-dissertation.social_employment_1.deflator_ipca_2024` AS d
ON
  r.ano = d.year
WHERE
  r.sigla_uf IN ('AC', 'AM', 'AP', 'MA', 'MT', 'PA', 'RO', 'RR', 'TO')
  AND r.vinculo_ativo_3112 = '1'
  AND r.valor_remuneracao_media IS NOT NULL
  AND r.valor_remuneracao_media > 0;
```

After running this script, I saved the resulting table in my personal Google BigQuery project. This intermediate table served as the basis for the next transformation step.

The **second query** constructs the wage-range categories and generates the agricultural versus non-agricultural sector dummy variable:

```{sql connection=}

SELECT
  ano AS year,
  id_municipio AS geocode,
  CASE
    WHEN LEFT(LPAD(cbo_unified, 6, '0'), 1) = '6' THEN 1
    ELSE 0
  END AS workers_agricultural,
CASE
  WHEN valor_remuneracao_media_2024 <= 1.00 * 1412.00 THEN 1
  WHEN valor_remuneracao_media_2024 > 1.00 * 1412.00 AND valor_remuneracao_media_2024 <= 2.00 * 1412.00 THEN 2
  WHEN valor_remuneracao_media_2024 > 2.00 * 1412.00 AND valor_remuneracao_media_2024 <= 3.00 * 1412.00 THEN 3
  WHEN valor_remuneracao_media_2024 > 3.00 * 1412.00 AND valor_remuneracao_media_2024 <= 4.00 * 1412.00 THEN 4
  WHEN valor_remuneracao_media_2024 > 4.00 * 1412.00 AND valor_remuneracao_media_2024 <= 5.00 * 1412.00 THEN 5
  ELSE 6
END AS wage_code,
  COUNT(*) AS workers_total
FROM
  `amz-data-dissertation.social_employment_1.rais_vinculos_deflat`
GROUP BY
  year,
  geocode,
  workers_agricultural,
  wage_code
ORDER BY
  year,
  geocode,
  workers_agricultural,
  wage_code;
```

Once executed, I saved this final processed table in my Google BigQuery project and subsequently exported it to my Google Cloud Storage environment. From there, I downloaded the dataset as a CSV file for use in the R data-analysis workflow.

### 2.2. Active Age Population (PIA)

#### 2.2.1. Understanding the data

Municipal population data for persons aged 14 years or older were obtained from the Brazilian Institute of Geography and Statistics (IBGE) Census for the years 2000, 2010, and 2022. These data correspond to the **Population in Active Age** **- População em Idade Ativa (PIA) in Portuguese** -, the number of individuals within each municipality who are old enough to be considered part of the potential labor force. This variable is fundamental for estimating labor-market indicators, such as the Economically Active Population - População Economicamente Ativa (PEA) -, when combined with employment and unemployment data.

#### 2.2.2. Import, and inspect the data

The data was obtained from the IBGE’s Sidra website. (<https://sidra.ibge.gov.br/pesquisa/censo-demografico/demografico-2022/amostra-nupcialidade-e-familia>)

#### 2.2.3. Clean and transform the data

Because the census data are available exclusively for 2000, 2010, and 2022, to construct a complete annual series for the entire study period, it is necessary to interpolate and extrapolate population values for all intermediate and adjacent years. I apply simple linear interpolation for the years between census observations and linear extrapolation for the years before and after census observations.

To do the interpolation, I use the `zoo` package to estimate the annual change between known census points, producing a smooth annual trajectory for each municipality from 2000 to 2022, as the equation shows:

$$
PIA_{it}
=
PIA_{i,t_1}
+
\frac{PIA_{i,t_2}-PIA_{i,t_1}}{t_2 - t_1}
\,(t - t_1),
$$

where $(t_1)$ and $(t_2)$ are the bounding census years (2000–2010 or 2010–2022).

For the years 1999, 2023 and 2024, I compute the extraploation trends manually. The value for 1999 is obtained by backward linear extrapolation:

$$
PIA_{i,1999}
=
PIA_{i,2000}
-
\frac{PIA_{i,2010} - PIA_{i,2000}}{10}.
$$

Future years are estimated by forward linear extrapolation based on the most recent census slope:

$$
PIA_{i,t}
=
PIA_{i,2022}
+
\frac{PIA_{i,2022} - PIA_{i,2010}}{12}
\,(t - 2022),
\qquad t \in \{2023, 2024\}.
$$

The code below shows all the interpolation and extrapolation transformations:

```{r}
library(dplyr)
library(tidyr)
library(zoo)  

population_pia <- population_pia %>%
  arrange(id, year) %>%
  group_by(id, state) %>%
  complete(year = 2000:2022)

population_pia <- population_pia %>%
  group_by(id, state) %>%
  mutate(total_interp = zoo::na.approx(total, x = year, na.rm = FALSE))

population_pia <- population_pia %>%
  arrange(id, year) %>%
  group_by(id, state) %>%
  complete(year = 1999:2024)

population_pia  <- population_pia %>%
  group_by(id, state) %>%
  mutate(
    first_year        = min(year[!is.na(total_interp)]),
    second_year       = sort(year[!is.na(total_interp)])[2],
    last_year         = max(year[!is.na(total_interp)]),
    second_last_year  = sort(year[!is.na(total_interp)], decreasing = TRUE)[2],
    
    first_val         = total_interp[year == first_year][1],
    second_val        = total_interp[year == second_year][1],
    last_val          = total_interp[year == last_year][1],
    second_last_val   = total_interp[year == second_last_year][1],
    
    slope_start = (second_val      - first_val)      / (second_year      - first_year),
    slope_end   = (last_val        - second_last_val)/ (last_year        - second_last_year),
    
    Total_full = dplyr::case_when(
      year < first_year ~ first_val + slope_start * (year - first_year),
      year > last_year  ~ last_val  + slope_end   * (year - last_year),
      TRUE              ~ total_interp
    )
  ) %>%
  ungroup() %>%
  select(state, id, year, total_interp = Total_full)


population_pia <- population_pia %>%
  mutate(total = total_interp) %>%
  select(-total_interp)
```

### 2.3. Labor and Wage Indicators

#### 2.3.1. Understanding the data

I combine the Employment and PIA datasets (geocode and year as keys) to normalize all employment variables by the population in active age (PIA). Normalizing employment by PIA allows for meaningful comparisons across municipalities with different population sizes and demographic structures.

To construct the labor indicators, I use employment data on (i) the total number of formal workers per municipality; (ii) the wage categories (six groups defined relative to the 2024 minimum wage), and (iii) the dummy variable identifying agricultural occupations.

I compute the ratios of each variable to the corresponding PIA value for each municipality-year. These normalized measures form the basis for analyzing the scale and composition of local labor markets across the Amazon region.

#### 2.3.2. Tidy and transform the data

First, I aggregate the labor indicators dataset by municipality, year, and wage code, summing the number of total workers and the number of agricultural workers. This step allow me to eliminate the agricultural dummy variable and work directly with counts of agricultural and total employment within each wage category:

```{r}
totals <- data_employment_rais %>%
  group_by(year, geocode, wage_code) %>%
  summarise(
    workers_total = sum(workers_total, na.rm = TRUE),
    .groups = "drop"
  )

totals_agricultural <- data_employment_rais %>%
  filter(workers_agricultural == 1) %>%     
  group_by(year, geocode, wage_code) %>%
  summarise(
    workers_agric = sum(workers_total, na.rm = TRUE),
    .groups = "drop"
  )

data_employment_rais_totals <- totals %>%
  left_join(totals_agricultural, by = c("year", "geocode", "wage_code")) %>%
  replace_na(list(workers_agric = 0L))
```

I also collapse the wage ranges previously constructed with the data I got from RAIS into three broader, substantive wage groups:

-   Low wage: less then 1 minimum wage

-   Middle wage: from 1 to 4 minimum wage

-   High wage: more then 4 minimum wage

This procedure reduces dimensionality while preserving substantive meaning, facilitating merging with other municipal-year datasets. I also constructed agricultural wage-range indicators, with the same wage-range grouping applied only to agricultural workers:

```{r}
rais_wage_groups <- data_employment_rais_totals %>%
  mutate(
    wage_group = case_when(
      wage_code %in% 1 ~ "low",
      wage_code %in% 2:4 ~ "mid",
      wage_code %in% 5:6 ~ "high"
    )
  ) %>%
  group_by(year, geocode, wage_group) %>%
  summarise(
    workers_group = sum(workers_total, na.rm = TRUE),
    .groups = "drop"
  )

data_employment_rais_wage_wide <- rais_wage_groups  %>%
  tidyr::pivot_wider(
    names_from = wage_group,
    values_from = workers_group,
    names_prefix = "workers_"
  ) %>%
  replace_na(list(
    workers_low = 0L,
    workers_mid = 0L,
    workers_high = 0L
  ))

rais_wage_groups_agri <- data_employment_rais_totals %>%
  mutate(
    wage_group = case_when(
      wage_code %in% 1 ~ "low",
      wage_code %in% 2:4 ~ "mid",
      wage_code %in% 5:6 ~ "high"
    )
  ) %>%
  group_by(year, geocode, wage_group) %>%
  summarise(
    workers_group = sum(workers_agric, na.rm = TRUE),
    .groups = "drop"
  )

data_employment_rais_wage_wide_agri <- rais_wage_groups_agri  %>%
  tidyr::pivot_wider(
    names_from = wage_group,
    values_from = workers_group,
    names_prefix = "workers_agri_"
  ) %>%
  replace_na(list(
    workers_low = 0L,
    workers_mid = 0L,
    workers_high = 0L
  ))
```

#### 2.3.3. Correct and standardize units

Using the ratio between the number of employees and the population in active age, I construct two types of indicators. The first set consists of **employment rates** by municipality, year, and wage group, calculated separately for all occupations and for agricultural occupations. The second set consists of **labor-market composition indicators**, which measure the proportion of workers in each wage group within a municipality-year, again distinguishing between all occupations and agricultural occupations.

##### 2.3.3.1. Employment Rates (Intensity Indicators)

The employment intensity indicators normalize the number of formal workers by the Population in Active Age (PIA). The indicator represent the density of formal jobs per active-age resident. The general form is:

$$
\text{EmploymentRate}_{it} = 
\frac{\text{Workers}_{it}}{PIA_{it}},
$$

where $(i)$ indexes municipalities and $(t)$ indexes years. I construct these intensity indicators for the following groups:

-   Total and wage-group employment rates:

    -   total employment rate (*emp_rate_total*)

    -   low wage employment rate (*emp_rate_low*)

    -   medium wage employment rate (*emp_rate_mid*)

    -   high wage employment rate (*emp_rate_high*)

-   Total agricultural wage-group employment rates:

    -   agricultural employment rate (*emp_rate_agric*)

    -   low wage agricultural employment rate (*emp_rate_agri_low*)

    -   medium wage agricultural employment rate (*emp_rate_agri_mid*)

    -   high wage agricultural employment rate (*emp_rate_agri_high*)

##### 2.3.3.2. Labor-Market Composition (Share Indicators)

Labor-market composition indicators express the proportion of workers in each wage group relative to total formal employment. These shares are calculated using a simple ratio:

$$
\text{Share_low_wage}_{it}
=
\frac{\text{Workers_low_wage}_{it}}{\text{Workers_total}_{it}},
$$

$$
\text{Share_mid_wage}_{it}
=
\frac{\text{Workers_mid_wage}_{it}}{\text{Workers_total}_{it}},
$$

$$
\text{Share_high_wage}_{it}
=
\frac{\text{Workers_high_wage}_{it}}{\text{Workers_total}_{it}}.
$$

The share of agricultural employment in total formal employment, as well as the distribution of wage groups within agricultural employment, are computed as follows:

$$
\text{Share_workers_agriculture}_{it}
=
\frac{\text{Workers_agricultural}_{it}}{\text{Workers_total}_{it}}.
$$

$$
\text{Share_low_wage_agricultural}_{it}
=
\frac{\text{Workers_low_wage_agricultural}_{it}}{\text{Workers_agricultural}_{it}},
$$

$$
\text{Share_mid_wage_agricultural}_{it}
=
\frac{\text{Workers_mid_wage_agricultural}_{it}}{\text{Workers_agricultural}_{it}},
$$

$$
\text{Share_high_wage_agricultural}_{it}
=
\frac{\text{Workers_high_wage_agricultural}_{it}}{\text{Workers_agricultural}_{it}}.
$$

Below, the code used to the computations:

```{r}
data_employment_rais_tidy <- data_employment_rais_totals %>%
  group_by(year, geocode) %>%
  summarise(
    workers_total = sum(workers_total, na.rm = TRUE),
    workers_agric = sum(workers_agric, na.rm = TRUE),
    .groups = "drop"
  )

data_employment_rais_full <- data_employment_rais_tidy %>%
  left_join(data_employment_rais_wage_wide,
            by = c("year", "geocode")) %>%
  left_join(data_employment_rais_wage_wide_agri,
            by = c("year", "geocode"))

data_employment_rais_full <- data_employment_rais_full %>%
  relocate(
    workers_low,
    workers_mid,
    workers_high,
    workers_agri_low,
    workers_agri_mid,
    workers_agri_high,
    .after = workers_agric
  )

data_population_pia %>% 
  filter(is.na(pia_total)) %>% 
  print(n=30)

data_population_pia <- data_population_pia %>%
  mutate(geocode = as.integer(id)) %>%
  select(geocode, year, pia_total = total)

data_labor_with_pia <- data_employment_rais_full %>%
  left_join(data_population_pia, by = c("year", "geocode"))

data_labor_with_pia <- data_labor_with_pia %>%
  mutate(
    emp_rate_total      = workers_total / pia_total,
    emp_rate_agric      = workers_agric / pia_total,
    emp_rate_low        = workers_low / pia_total,
    emp_rate_mid        = workers_mid / pia_total,
    emp_rate_high       = workers_high / pia_total,
    emp_rate_agri_low   = workers_agri_low / pia_total,
    emp_rate_agri_mid   = workers_agri_mid / pia_total,
    emp_rate_agri_high  = workers_agri_high / pia_total,
      )

data_labor_with_pia <- data_labor_with_pia %>%
  mutate(
    share_low_workers   = workers_low  / workers_total,
    share_mid_workers   = workers_mid  / workers_total,
    share_high_workers  = workers_high / workers_total,
    share_agri_low_workers  = workers_agri_low  / workers_agric,
    share_agri_mid_workers  = workers_agri_mid  / workers_agric,
    share_agri_high_workers = workers_agri_high / workers_agric,
    share_agri_workers  = workers_agric / workers_total
  )
```

**Observation:** The municipality of Mojuí dos Campos (PA) — geocode 1504752 — contains missing values for all years except 2022. This occurs because Mojuí dos Campos was formerly a district of Santarém (PA) and became an independent municipality only in 2013. The missing observations were retained at this stage and will be addressed later during the construction of the complete dataset.

### 2.4. Cash transfer (Bolsa Família program)

#### 2.4.1. Understand the data

The cash transfer data used in this study comes from administrative records of the Bolsa Família program published by Brazil’s **Ministério do Desenvolvimento e Assistência Social, Família e Combate à Fome (MDS)** in the **Dados Abertos (Open Data)** portal of the Federal Government of Brazil (last updated in March 2026, available at <https://dados.gov.br/dados/conjuntos-dados/bolsa-familia>).

The raw data is organized as monthly municipality-level files (one file per year), with CSV files for each year from 2004 (the first year of the policy) to 2026. Because my final dataset is a municipality-year panel, I select a **reference month** to represent each year. I conduct stability checks on candidate months (see below), and end up with **February** and **June** as alternative reference months for robustness. I also restrict the panel to **2004–2019**, excluding 2020 onward due to the COVID-19 period and program substitution dynamics that generate structural discontinuities in the Bolsa Família series.

The standard structure of the raw files includes:

-   `ibge`: IBGE municipality code
-   `anomes`: reference year-month (format `YYYYMM`)
-   `qtd_familias_beneficiarias_bolsa_familia`: number of beneficiary families in the municipality-month
-   `valor_repassado_bolsa_familia`: total value transferred to beneficiary families in the municipality-month

These variables allow the construction of three complementary dimensions of program intensity:

i.  **Coverage**: number of beneficiary families

ii. **Generosity**: average transfer per beneficiary family (total transfers / beneficiary families)

iii. **State effort**: total transfers (in constant prices)

#### 2.4.2. Import and inspect the data

I store the raw files locally as CSV files with the naming convention `bf-YYYY.txt`. I read all files into R using a function that applies consistent parsing rules and adds a `source_file` variable to track the origin of each observation.

```{r}

bf_path <- "_data/raw_mds_bolsa_familia"

bf_files <- list.files(
  path = bf_path,
  pattern = "^bf-\\d{4}\\.txt$",
  full.names = TRUE
)

read_bf <- function(file) {
  read_csv(
    file,
    col_types = cols(
      ibge = col_double(),
      anomes = col_double(),
      qtd_familias_beneficiarias_bolsa_familia = col_double(),
      valor_repassado_bolsa_familia = col_double()
    )
  ) %>%
    mutate(source_file = basename(file))
}

data_bf_all <- map_dfr(bf_files, read_bf)
```

I apply a set of integrity checks to ensure that the two key variables—beneficiary families and total transfers—are consistently recorded across yearly files. These include missingness, zeros, negative values, logical inconsistencies (beneficiary families with zero transfers), and duplicate municipality-month keys.

```{r}
bf_qa_summary <- data_bf_all %>%
  group_by(source_file) %>%
  summarise(
    n_rows = n(),
    na_families = sum(is.na(qtd_familias_beneficiarias_bolsa_familia)),
    na_transfers = sum(is.na(valor_repassado_bolsa_familia)),
    zero_families = sum(qtd_familias_beneficiarias_bolsa_familia == 0, na.rm = TRUE),
    zero_transfers = sum(valor_repassado_bolsa_familia == 0, na.rm = TRUE),
    neg_families = sum(qtd_familias_beneficiarias_bolsa_familia < 0, na.rm = TRUE),
    neg_transfers = sum(valor_repassado_bolsa_familia < 0, na.rm = TRUE),
    families_pos_transfers_zero = sum(
      qtd_familias_beneficiarias_bolsa_familia > 0 & valor_repassado_bolsa_familia == 0,
      na.rm = TRUE
    ),
    duplicated_ibge_anomes = sum(duplicated(paste(ibge, anomes))),
    .groups = "drop"
  )

bf_qa_summary
```

I also compute distributional diagnostics for the implied average transfer per family (`valor_repassado_bolsa_familia / qtd_familias_beneficiarias_bolsa familia`) as a way to detect anomalous tails that may reflect reporting artifacts or program regime changes.

```{r}
bf_transfer_dist <- data_bf_all %>%
  mutate(
    avg_transfer = if_else(
      qtd_familias_beneficiarias_bolsa_familia > 0,
      valor_repassado_bolsa_familia / qtd_familias_beneficiarias_bolsa_familia,
      NA_real_
    )
  ) %>%
  group_by(source_file) %>%
  summarise(
    p1 = quantile(avg_transfer, 0.01, na.rm = TRUE),
    median = median(avg_transfer, na.rm = TRUE),
    p99 = quantile(avg_transfer, 0.99, na.rm = TRUE),
    max = max(avg_transfer, na.rm = TRUE),
    .groups = "drop"
  )

bf_transfer_dist
```

To select the reference months, I conduct **stability checks** (among others, see the socia data cleaning procedure in the respective script [`02-data-cleaning-populational-social.R`](_scripts/02-data-cleaning-populational-social.R)), comparing the number of beneficiary families and the average transfer per family in candidate months (February, June, July, September, and November) with their adjacent months. I do not chose January neither December because they will lack some adjacent months for the stability checks in the first and final years of the series (2004 and 2019).

The goal is to identify months that do not exhibit anomalous spikes or drops that may reflect reporting artifacts or program regime changes. I end up with June as the primary reference month for the panel, with February as an alternative reference month for robustness checks.

```{r}
candidate_months <- c(2, 6, 7, 9, 11)

library(dplyr)
library(tidyr)

bf_m <- bf_all %>%
  filter(anomes >= 200401 & anomes <= 201912) %>%
  mutate(
    year  = as.integer(substr(anomes, 1, 4)),
    month = as.integer(substr(anomes, 5, 6)),
    valor_por_familia = if_else(
      qtd_familias_beneficiarias_bolsa_familia > 0,
      valor_repassado_bolsa_familia /
        qtd_familias_beneficiarias_bolsa_familia,
      NA_real_
    )
  )

tail(bf_m)

stability <- bf_m %>%
  filter(month %in% c(candidate_months,
                      candidate_months - 1,
                      candidate_months + 1)) %>%
  group_by(ibge, year) %>%
  arrange(month) %>%
  mutate(
    fam_lag = lag(qtd_familias_beneficiarias_bolsa_familia),
    fam_lead = lead(qtd_familias_beneficiarias_bolsa_familia),
    val_lag = lag(valor_por_familia),
    val_lead = lead(valor_por_familia)
  ) %>%
  filter(month %in% candidate_months) %>%
  summarise(
    month = first(month),
    fam_instability =
      mean(abs(qtd_familias_beneficiarias_bolsa_familia - fam_lag) /
             fam_lag, na.rm = TRUE),
    val_instability =
      mean(abs(valor_por_familia - val_lag) /
             val_lag, na.rm = TRUE),
    .groups = "drop"
  )
print(stability, n = 22)
```

#### 2.4.3. Clean and transform the data

To build a municipality-year panel, I extract the year and month from `anomes`, filter to the analysis window (2004–2019), and select February and June as reference months. I then construct:

-   `bf_beneficiary_families`: coverage (number of beneficiary families)
-   `bf_total_transfers_brl`: state effort (nominal BRL)
-   `bf_avg_transfer_brl`: generosity (nominal BRL per beneficiary family)
-   `bf_reference_month`: indicate the reference month (Feb or Jun)

Coverage is measured by the number of beneficiary families, while generosity is calculated as the average transfer per family. State effort is represented by the total transfers. The resulting panel is sorted by reference month, year, and municipality geocode.

```{r}
data_bf_panel <- data_bf_all %>%
  mutate(
    year  = as.integer(str_sub(as.character(anomes), 1, 4)),
    month = as.integer(str_sub(as.character(anomes), 5, 6))
  ) %>%
  filter(year >= 2004, year <= 2019, month %in% c(2, 6)) %>%
  transmute(
    geocode = as.character(as.integer(ibge)),
    year = year,
    bf_ref_month = recode(month, `2` = "Feb", `6` = "Jun"),
    bf_beneficiary_families = qtd_familias_beneficiarias_bolsa_familia,
    bf_total_transfers_brl = valor_repassado_bolsa_familia,
    bf_avg_transfer_brl = if_else(
      qtd_familias_beneficiarias_bolsa_familia > 0,
      valor_repassado_bolsa_familia / qtd_familias_beneficiarias_bolsa_familia,
      NA_real_
    )
  ) %>%
  arrange(bf_ref_month, year, geocode)

data_bf_panel %>% count(bf_ref_month, year)
```

#### 2.4.4. Correct and standardize units

##### 2.4.4.1. Deflation to constant prices

To compare monetary values over time, I deflate Bolsa Família transfers using an IPCA-based deflator with **2024 as the base year** (deflator multiplier). I apply the deflator to both total transfers and the average transfer per family.

The *deflator data is the same that I use to deflate labor and wage indicators*. See @sec-deflation for details on the construction of the deflator table.

```{r}
deflator_path <- "_data/deflator_ipca_2024.csv"

data_ipca_deflator <- read_csv(
  deflator_path,
  col_types = cols(
    year = col_integer(),
    ipca_index = col_double(),
    ipca_deflator = col_double()
  )
) %>%
  select(year, ipca_deflator)

data_bf_panel_real <- data_bf_panel %>%
  left_join(data_ipca_deflator, by = "year") %>%
  mutate(
    bf_total_transfers_brl_2024 = bf_total_transfers_brl * ipca_deflator,
    bf_avg_transfer_brl_2024    = bf_avg_transfer_brl * ipca_deflator
  )

data_bf_panel_real %>%
  summarise(
    n = n(),
    missing_deflator = sum(is.na(ipca_deflator)),
    min_year = min(year),
    max_year = max(year)
  )
```

Finally, I compute a simple annual diagnostic summary to confirm that the transformation does not introduce unexpected missingness patterns.

```{r}
data_bf_panel_real %>%
  group_by(bf_ref_month, year) %>%
  summarise(
    n = n(),
    na_avg_transfer = mean(is.na(bf_avg_transfer_brl_2024)),
    zero_total_transfers = mean(bf_total_transfers_brl_2024 == 0, na.rm = TRUE),
    .groups = "drop"
  )
```

#### 2.4.5. Check and validate the data

I conduct a final set of checks to confirm that the the dataset is consistent with known patterns of the Bolsa Família program. I examine the temporal trends of total transfers, beneficiary families, and average transfer per family across the two reference months to ensure that they align with expected trajectories based on program history and previous research.

The plots of the annual trajectories of the three key indicators are consistent with the known expansion and stabilization phases of the Bolsa Família program over the 2004–2019 period as shown by @souza2022 .

```{r}
bf_effort_yearly <- data_bf_panel_deflate %>%
  group_by(bf_ref_month, year) %>%
  summarise(
    real_total_transfers =
      sum(bf_total_transfers_2024, na.rm = TRUE),
    total_beneficiary_families =
      sum(bf_beneficiary_families, na.rm = TRUE),
    avg_transfer_2024 =
      real_total_transfers / total_beneficiary_families,
    .groups = "drop"
  )

ggplot(bf_effort_yearly,
       aes(x = year,
           y = real_total_transfers,
           linetype = bf_ref_month)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Total real Bolsa Família transfers (2024 BRL)",
    x = "Year",
    y = "Total transfers (2024 BRL)",
    linetype = "Reference month"
  ) +
  theme_minimal()

ggplot(bf_effort_yearly,
       aes(x = year,
           y = total_beneficiary_families,
           linetype = bf_ref_month)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Total Bolsa Família beneficiary families",
    x = "Year",
    y = "Total beneficiary families",
    linetype = "Reference month"
  ) +
  theme_minimal()

ggplot(bf_effort_yearly,
       aes(x = year,
           y = avg_transfer_2024,
           linetype = bf_ref_month)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Bolsa Família transfer per beneficiary family (2024 BRL)"
    subtitle = "Yearly averages using February and June as reference months",
    x = "Year",
    y = "Average transfer per family (2024 BRL)",
    linetype = "Reference month"
  ) +
  theme_minimal()

```

#### 2.4.6. Standardize units and construct indicators

To enable comparisons across municipalities with different population sizes, I standardize the units of the Bolsa Família indicators using the following complementary demographic and policy data:

*Economically Active Population / População Economicamente Ativa (PEA)*: the total number of individuals participating in the labor market (either employed or actively seeking employment). It allows the scaling of social transfers relative to the size of the **local labor force**.

I obtained PEA data in each municipality-year from the Brazilian population censuses (IBGE) for the years 2000, 2010, and 2022, and interpolate it to construct a complete annual panel.

*Bolsa Família municipal quota*: defined by the federal government based on official estimates of the number of poor families (such as PNAD surveys). As such, represents the maximum number of beneficiary families that can be served by the program in a given municipality and serves as a **policy target for coverage**, indicating the intended scale of the program in each locality.

Mncipal quotas has been updated overtime as follows:

-   **Up to March 2006:**\
    Estimates based on the **2001 PNAD (Pesquisa Nacional por Amostra de Domicílios)**

-   **April 2006 to April 2009:**\
    Estimates based on the **2004 PNAD**

-   **May 2009 to April 2012:**\
    Estimates based on the **Mapa de Pobreza** (Poverty Map), which incorporates small-area estimation techniques to improve geographic precision

-   **From May 2012 onward:**\
    Estimates based on the **2010 PNAD**

I obtained the municipal quota data using the federal transparency portal **Busca LAI**, which gather formal requests of data under Brazil’s **Access to Information Law / Lei de Acesso à Informação (LAI)**. The request was submitted to the Ministry of Development and Social Assistance, Family and Fight against Hunger (MDS) (Protocol number 71003.017363/2022-16, available at <https://buscalai.cgu.gov.br/busca/4388082>, last acess in April 2026).

*Informality rate:* I also gathered data on the proportion of employed individuals in informal work arrangements, including workers without a formal contract, self-employed individuals without CNPJ registration, and unpaid family workers.

The **employed population (população ocupada)** includes individuals who performed any economic activity during the reference period, either for wages, profit, or family production.

Brazilian labor statistics classify workers by **position in the occupation (posição na ocupação)**. Informality is not directly reported in the Demographic Census and must be constructed from these occupational categories.

Informal workers include:

-   Private employees **without a formal contract**
-   Domestic workers **without a formal contract**
-   Self-employed **without registered business**
-   Employers **without registered business**
-   Unpaid family workers

Formal workers include:

-   Private employees **with a formal contract**
-   Domestic workers **with a formal contract**
-   Public sector employees and military
-   Self-employed **with registered business**
-   Employers **with registered business**

##### 2.4.6.1. Economically Active Population (PEA) data

I construct a balanced annual panel for the economically active population (PEA) and generate interpolated values to fill gaps between census years. I apply **linear interpolation** within each municipality using the available census observations (2000, 2010, and 2022). The interpolated variable is computed by linearly connecting observed values across time, producing a smooth annual series of the economically active population for each municipality.

Interpolation was performed within municipalities, preserving the temporal structure of each unit and avoiding cross-sectional assumptions. The resulting dataset provides a complete municipality-year panel (2000–2022) of PEA values, combining observed census data and interpolated estimates, and is used in subsequent stages to standardize social and labor market indicators.

Before performing interpolation, it is necessary to address cases of municipal boundary changes that generate missing observations in earlier years. The only substantively relevant case in the sample is **Mojuí dos Campos (PA)** (IBGE code **1504752**), which was created in **2013** through the territorial dismemberment of **Santarém (PA)** (code **1506807**). As a result, Mojuí dos Campos does not have census observations for 2000 and 2010, and only appears in the data from 2022.

To address it, I first calculate the share of the combined 2022 PEA of Santarém and Mojuí dos Campos that corresponds to Mojuí dos Campos. I then use this proportion as a fixed allocation rule to reconstruct Mojuí dos Campos’s historical values for 2000 and 2010. Correspondingly, Santarém’s pre-2013 PEA values were reduced by the complementary share, so that the historical total of the original territory was redistributed between the two municipalities in a consistent way.

This procedure produces a territorially harmonized series in which Mojuí dos Campos and Santarém are treated as analytically comparable units over time. The adjustment assumes that the relative demographic weight of the two territories observed in 2022 approximates their relative distribution in earlier years. Although this assumption cannot be verified directly for the pre-creation period, it is a reasonable approximation for backward harmonization and is preferable to leaving Mojuí dos Campos missing or inconsistently defined in the panel.

In the deforestation dataset @ref(#sec-deforestation), this harmonization is already implemented, and the procedure described here ensures consistency between socioeconomic and environmental variables.

```{r}
population_pea_2022 <- population_pea %>%
  filter(year == 2022) %>%
  select(geocode, pea)

pea_pair <- population_pea_2022 %>%
  filter(geocode %in% c(1506807, 1504752))

share_mojui <- pea_pair %>%
  summarise(
    share = pea[geocode == 1504752] /
      (pea[geocode == 1504752] + pea[geocode == 1506807])
  ) %>%
  pull(share)

population_pea_mj_adjusted <- population_pea %>%
  mutate(geocode = as.numeric(geocode)) %>%
  bind_rows(
    tibble(
      geocode = 1504752,
      pea = NA_real_,
      year = c(2000, 2010)
    )
  ) %>%
  distinct(geocode, year, .keep_all = TRUE) %>%
  left_join(
    population_pea %>%
      mutate(geocode = as.numeric(geocode)) %>%
      filter(geocode == 1506807) %>%
      select(year, pea_santarem = pea),
    by = "year"
  ) %>%
  mutate(
    pea = case_when(
      geocode == 1504752 & year < 2013 ~ pea_santarem * share_mojui,
      geocode == 1506807 & year < 2013 ~ pea_santarem * (1 - share_mojui),
      TRUE ~ pea
    )
  ) %>%
  select(-pea_santarem)


population_pea_panel <- population_pea_mj_adjusted %>%
  mutate(
    geocode = as.numeric(geocode),
    year = as.integer(year)
  ) %>%
  group_by(geocode, year) %>%
  summarise(pea = first(pea), .groups = "drop") %>%
  group_by(geocode) %>%
  complete(year = 2000:2022) %>%
  arrange(geocode, year) %>%
  mutate(
    pea_interp = zoo::na.approx(pea, x = year, na.rm = FALSE)
  ) %>%
  ungroup()

```

##### 2.4.6.2. Bolsa Família Municipal Quota data

I reshaped the municipal quota dataset into a municipality-year panel and then merged it with the deflated Bolsa Família panel. Before merging, I harmonized the municipality identifiers across the two sources.

Before the merge, I identified an inconsistency in the geographic identifiers: the quota dataset used **7-digit IBGE municipality codes**, while the deflated Bolsa Família panel used **6-digit codes**. To make the two datasets compatible, I dropped the last digit of the 7-digit quota geocode, creating a harmonized 6-digit identifier (`geocode_6`). I then created the same harmonized field in the Bolsa Família panel and merged the two datasets by **municipality** and **year**.

```{r}
data_bf_quota <- read_xlsx("_data/raw_mds_bolsa_familia/bf-quotas.xlsx")

data_bf_quota_panel <- data_bf_quota %>%
  pivot_longer(
    cols = starts_with ("quota_"),
    names_to = "year",
    values_to = "bf_quota"
  ) %>% 
  mutate(
    year = str_remove(year, "quota_") |> as.integer(),
    geocode = as.integer(geocode),
    bf_quota = as.numeric(bf_quota)
  ) %>%
  arrange(geocode, year)
  
data_bf_quota_panel <- data_bf_quota_panel %>%
  mutate(
    geocode_6 = geocode %/% 10
  )

data_bf_panel_deflate <- data_bf_panel_deflate %>%
  mutate(
    geocode_6 = geocode
  )

data_bf_panel_final <- data_bf_panel_deflate %>% 
  left_join(data_bf_quota_panel, by = c("geocode_6", "year"))

```

##### 2.4.6.3. Final Bolsa Família dataset

To construct the final Bolsa Família, I restrict to **June**, which serve as the main reference period in the analysis (I repeat after the same procedures filtering for **Feb**).

I standardized variable names to improve clarity and consistency. - defined **bf_families_n** as the number of beneficiary families,\
- defined **bf_transfers_total_brl_2024** as the total transfers (in 2024 BRL), and\
- defined **bf_transfers_avg_brl_2024** as the average transfer per family (in 2024 BRL).

Next, I merged the dataset with the **economically active population (PEA) panel**, matching by municipality and year. From the merged dataset, I constructed three key indicators:

-   **Coverage (quota ratio):**

$$
Coverage_{it} = \frac{BeneficiaryFamilies_{it}}{Quota_{it}}
$$

-   **Effort per economically active population (PEA):**

$$
Effort^{PEA}_{it} = \frac{TotalTransfers_{it}}{PEA_{it}}
$$

-   **Effort per quota:**

$$
Effort^{Quota}_{it} = \frac{TotalTransfers_{it}}{Quota_{it}}
$$

These indicators jointly capture **coverage**, **generosity**, and **fiscal intensity** of the program at the municipal level.

```{r}
outcome_bf <- data_bf_panel_final %>%
  filter(bf_ref_month == "Jun") %>%
  rename(
    geocode = geocode.y,
    bf_families_n = bf_beneficiary_families,
    bf_transfers_total_brl_2024 = bf_total_transfers_2024,
    bf_transfers_avg_brl_2024 = bf_avg_transfer_2024
  ) %>%
  select(
    geocode,
    year,
    bf_families_n,
    bf_transfers_total_brl_2024,
    # Generosity
    bf_transfers_avg_brl_2024,
    bf_quota
  ) %>%
  left_join(
    data_population_pea_panel %>% 
    select(geocode, year, pea), 
    by = c("geocode", "year")) %>%
  mutate(
    # Coverage (ratio)
    bf_families_quota_ratio = if_else(
      !is.na(bf_quota) & bf_quota > 0,
      bf_families_n / bf_quota,
      NA_real_
    ),
    # Effort per PEA (in 2024 BRL)
    bf_transfers_pea_brl_2024 = if_else(
      !is.na(pea) & pea > 0,
      bf_transfers_total_brl_2024 / pea,
      NA_real_
    ),

    # Effort (in 2024 BRL)
    bf_transfers_quota_brl_2024 = if_else(
      !is.na(bf_quota) & bf_quota > 0,
      bf_transfers_total_brl_2024 / bf_quota,
      NA_real_
    )
  ) %>% 
  arrange(geocode, 
          year)
```

## 3. Control variables

### 3.1. Labor market informality

#### 3.1.1. Understand the data

The labor market informality data used in this study comes from the **Brazilian Demographic Census** (*Censo Demográfico*), conducted by the **Instituto Brasileiro de Geografia e Estatística (IBGE)**. I use the census waves of **2000**, **2010**, and **2022**, which are the three waves available within the study's time frame.

The IBGE operationalizes labor market informality based on the worker's posição na ocupação (position in occupation), a categorical variable collected in the Census questionnaire. The following categories are classified as informal: private-sector employees without a signed work permit (carteira de trabalho assinada); domestic workers without a signed work permit; employers not registered with the CNPJ (the Brazilian tax registry for firms); own-account workers (trabalhadores por conta própria) without CNPJ registration; and unpaid contributing family workers (trabalhadores familiares auxiliares). This definition follows the ILO standard for measuring informal employment and is the same criterion IBGE applies in its official informality rate releases via the PNAD Contínua. Formally employed workers — those with a signed work permit in the private sector, public servants, military personnel, and registered employers — constitute the complement. The informality rate is therefore a measure of the prevalence of unprotected, unregistered labor relationships within the occupied population of each municipality, capturing both employment vulnerability and the local reach of labor market institutions.

The Census provides municipality-level counts of the working-age population classified as employed in either the formal or informal sector. The variable of interest is the **informality rate**: the share of the occupied population working in informal arrangements, defined by the absence of a signed work permit (*carteira de trabalho assinada*) or equivalent formal employment bond.

Because the Census is conducted at decennial intervals (with the addition of the 2022 wave), data is only available for three benchmark years across the 2000–2022 period. I linearly interpolate the informality rate between census years to construct a continuous annual panel, under the assumption that structural change in local labor markets follows a gradual trajectory between observation points.

To preserve temporal consistency in the municipal panel, I applied a backward harmonization procedure. I inserted observations for Mojuí dos Campos in 2000 and 2010 and assigned to them the same formal and informal labor-market shares observed for Santarém in those years. This procedure assumes that, before administrative separation, the labor-market structure of the future Mojuí dos Campos territory was similar to that of Santarém. Although this is an approximation, it avoids treating the absence of Mojuí dos Campos before 2013 as a substantive missing-data pattern and ensures consistency with the municipality-year structure used in the rest of the analysis.

#### 3.1.2. Import and inspect the data

I import three separate Excel files, one per census wave. Each file contains municipality identifiers (`geocode`), municipality names (`munic`), and counts of formally and informally employed workers (`Formal`, `Informal`). I add a `year` variable to each and strip non-numeric characters from the geocode field to ensure consistent integer coding.

```{r}
population_informal_2000 <- read_excel("_data/raw_censo_labor_population/informality-2000.xlsx")
population_informal_2010 <- read_excel("_data/raw_censo_labor_population/informality-2010.xlsx")
population_informal_2022 <- read_excel("_data/raw_censo_labor_population/informality-2022.xlsx")

population_informal_2022 <- population_informal_2022 %>%
  mutate(year    = 2022,
         geocode = as.integer(gsub("[^0-9]", "", geocode)))

population_informal_2010 <- population_informal_2010 %>%
  mutate(year    = 2010,
         geocode = as.integer(gsub("[^0-9]", "", geocode)))

population_informal_2000 <- population_informal_2000 %>%
  mutate(year    = 2000,
         geocode = as.integer(gsub("[^0-9]", "", geocode)))

population_informal <- bind_rows(
  population_informal_2000,
  population_informal_2010,
  population_informal_2022
) %>%
  arrange(geocode, year)
```

I check for duplicate municipality-year entries, which would indicate an import or reshaping error, and identify municipalities that are present in fewer than three census waves — typically units created by municipal splits after 2000.

```{r}
# Check for duplicates
population_informal %>%
  count(geocode, year) %>%
  filter(n > 1)

# Identify municipalities with incomplete census coverage
population_informal %>%
  mutate(geocode = as.numeric(geocode)) %>%
  filter(geocode != 1) %>%
  distinct(geocode, year) %>%
  count(geocode) %>%
  filter(n < 3)
```

#### 3.1.3. Handle the Mojuí dos Campos municipality split

As with the GDP and population variables, the creation of **Mojuí dos Campos** (IBGE geocode `1504752`) from **Santarém** (geocode `1506807`) in 2013 requires a backcasting adjustment to maintain panel consistency. Mojuí dos Campos has no Census records for 2000 or 2010, since it did not exist as an independent administrative unit at those reference dates.

I address this by inserting placeholder rows for Mojuí dos Campos in 2000 and 2010 with missing values, then imputing those values directly from Santarém's observed formal and informal employment counts for the same years. This approach differs slightly from the GDP backcasting procedure (see Section 3.2.4), where proportional shares were applied: here, because the Census already records employment by municipality and the pre-split territory of Mojuí was administratively part of Santarém, I use Santarém's raw counts as the basis for imputation, under the assumption that the two territories' labor market structure was indistinguishable prior to the split. The interpolation step that follows (Section 3.1.4) then produces smooth trajectories for both municipalities between the census anchor points.

```{r}
population_informal_mj_adjusted <- population_informal %>%
  mutate(geocode = as.numeric(geocode)) %>%
  bind_rows(
    tibble(
      geocode  = 1504752,
      munic    = "Mojuí dos Campos (PA)",
      Formal   = NA_real_,
      Informal = NA_real_,
      year     = c(2000, 2010)
    )
  ) %>%
  distinct(geocode, year, .keep_all = TRUE) %>%
  left_join(
    population_informal %>%
      mutate(geocode = as.numeric(geocode)) %>%
      filter(geocode == 1506807) %>%
      select(year,
             informal_santarem = Informal,
             formal_santarem   = Formal),
    by = "year"
  ) %>%
  mutate(
    Informal = case_when(
      geocode == 1504752 & year < 2013 ~ informal_santarem,
      TRUE ~ Informal
    ),
    Formal = case_when(
      geocode == 1504752 & year < 2013 ~ formal_santarem,
      TRUE ~ Formal
    )
  ) %>%
  select(-informal_santarem, -formal_santarem)
```

#### 3.1.4. Construct the informality rate and interpolate

I compute the informality rate as the share of informal workers in the total occupied population (formal + informal) for each municipality-year observation:

$$
\text{Informality}_{it} = \frac{\text{Informal}_{it}}{\text{Formal}_{it} + \text{Informal}_{it}}
$$

where $\text{Informal}_{it}$ and $\text{Formal}_{it}$ are the counts of informally and formally employed workers in municipality $i$ at census year $t$.

I then expand the panel to cover all years from 2000 to 2022 using `complete()` and apply linear interpolation via `zoo::na.approx()` to fill non-census years. The formal rate is derived as the complement:

$$
\text{Formal}_{it'} = 1 - \text{Informality}_{it'}, \qquad \forall \ t' \notin \{2000, 2010, 2022\}
$$

Municipality names are propagated across the expanded rows using a bidirectional fill.

```{r}
population_informal_panel <- population_informal_mj_adjusted %>%
  mutate(geocode = as.numeric(geocode),
         year    = as.integer(year)) %>%
  group_by(geocode, year) %>%
  summarise(
    munic    = first(munic),
    informal = first(Informal),
    .groups  = "drop"
  ) %>%
  group_by(geocode) %>%
  complete(year = 2000:2022) %>%
  arrange(geocode, year) %>%
  fill(munic, .direction = "downup") %>%
  mutate(
    informal_interp = zoo::na.approx(informal, x = year, na.rm = FALSE),
    formal_interp   = 100 - informal_interp
  ) %>%
  ungroup()
```

#### 3.1.5. Check and validate the data

I inspect the interpolated series visually for Mojuí dos Campos and Santarém to verify that the trajectories are plausible and internally consistent. I also confirm that no municipality has a constant informality rate across all years (which would suggest a failed interpolation) and that the series is bounded between 0 and 100.

```{r}
# Visual inspection for Mojuí dos Campos and Santarém
ggplot(filter(population_informal_panel, geocode == 1504752),
       aes(x = year, y = informal_interp)) +
  geom_line(linewidth = 1.2, color = "salmon") +
  geom_point() +
  theme_minimal() +
  labs(title = "Informality rate — Mojuí dos Campos (1504752)",
       x = "Year", y = "Informality rate (%)")

ggplot(filter(population_informal_panel, geocode == 1506807),
       aes(x = year, y = informal_interp)) +
  geom_line(linewidth = 1.2, color = "red") +
  geom_point() +
  theme_minimal() +
  labs(title = "Informality rate — Santarém (1506807)",
       x = "Year", y = "Informality rate (%)")

# Bounds check
population_informal_panel %>%
  summarise(
    min_informal = min(informal_interp, na.rm = TRUE),
    max_informal = max(informal_interp, na.rm = TRUE),
    n_na         = sum(is.na(informal_interp))
  )
```

The final dataset retains only the geocode, year, and interpolated informality rate, saved in CSV, Excel, and RDS formats for downstream merging.

```{r}
data_population_informal_panel <- population_informal_panel %>%
  select(geocode, year, informal_interp) %>%
  rename(informal = informal_interp)

data_population_informal_panel %>% write_csv("_data/outcome_population_informality.csv")
data_population_informal_panel %>% write_xlsx("_data/outcome_population_informality.xlsx")
data_population_informal_panel %>% write_rds("_data/outcome_population_informality.rds")
```


### 3.2. Municipal GDP per capita

#### 3.2.1. Understand the data

Municipal GDP data comes from the **Instituto Brasileiro de Geografia e Estatística (IBGE)**, which publishes the *Produto Interno Bruto dos Municípios* series annually. The variable of interest is GDP at current prices, expressed in thousands of Brazilian Reais (*Produto Interno Bruto a preços correntes — Mil Reais*). The dataset covers Brazilian municipalities from 2002 to 2023.

Population data, required to compute GDP per capita, comes from the **2000, 2010 and 2022 Demographic Censuses**, which provide headcount population for those benchmark years. Since census data is only available at decennial intervals (plus the 2022 wave), I interpolate population linearly across non-census years to construct a continuous annual panel, following the same procedure applied to other socioeconomic variables in this dataset.

To allow temporal comparability, all GDP values are **deflated to 2024 Brazilian Reais** using the **IPCA** national consumer price index (IBGE). The deflator is the same used for wage and Bolsa Família transfer variables throughout this appendix (see @sec-deflation for details on its construction).

#### 3.2.2. Import and inspect the data

The raw GDP file requires skipping three header rows and setting a maximum row count to exclude metadata appended at the bottom of the file by IBGE. Missing value codes used by IBGE (`"..."` and `"-"`) are declared explicitly at import so that they are correctly parsed as `NA` rather than treated as strings.

```{r}
data_ipca_deflator <- read_csv(
  "_data/deflator_ipca_2024.csv",
  col_types = cols(
    year          = col_integer(),
    ipca_index    = col_double(),
    ipca_deflator = col_double()
  )
) %>%
  select(year, ipca_deflator)

data_pib_munic     <- read_csv(
  "_data/raw_ibge_pib/pib_munic.csv",
  skip  = 3,
  n_max = 5570,
  na    = c("", "NA", "...", "-")
)

data_pop_2000_2010 <- read_excel("_data/raw_ibge_pib/pop_censo_2010_2000.xlsx")
data_pop_2022      <- read_excel("_data/raw_ibge_pib/pop_censo_2022.xlsx")
```

#### 3.2.3. Clean and transform the data

**Population panel.** I reshape the 2000/2010 census data and the 2022 census data from wide to long format and bind them into a single panel. I then expand the panel to include all years from 2000 to 2022 using `complete()` and linearly interpolate population values between census years using `zoo::na.approx()`. Municipality names are filled in both directions to avoid `NA`s arising from the expansion step.

```{r}
data_pop_2000_2010_long <- data_pop_2000_2010 %>%
  mutate(across(c(`2000`, `2010`), ~ parse_number(as.character(.)))) %>%
  pivot_longer(cols      = c(`2000`, `2010`),
               names_to  = "year",
               values_to = "population") %>%
  mutate(year = as.integer(year), geocode = as.character(geocode)) %>%
  select(geocode, municipality, year, population)

data_pop_2022_long <- data_pop_2022 %>%
  mutate(year     = 2022L,
         geocode  = as.character(geocode),
         population = as.numeric(population)) %>%
  select(geocode, municipality, year, population)

data_population_panel <- bind_rows(data_pop_2000_2010_long,
                                   data_pop_2022_long) %>%
  arrange(geocode, year) %>%
  group_by(geocode) %>%
  complete(year = 2000:2022) %>%
  arrange(year) %>%
  mutate(
    population = zoo::na.approx(population, x = year, na.rm = FALSE)
  ) %>%
  fill(municipality, .direction = "downup") %>%
  ungroup() %>%
  select(geocode, year, municipality, population)
```

**GDP panel.** I reshape the wide GDP file to long format, converting year columns (2002–2023) to an integer variable and GDP values to numeric. I convert GDP from thousands to full Reais by multiplying by 1,000. I then join the IPCA deflator and deflate all values to 2024 Reais:

$$
GDP^{2024}_{it} = GDP^{nominal}_{it} \times \text{Deflator}(t), \qquad \text{where} \quad \text{Deflator}(t) = \frac{IPCA_{2024}}{IPCA_t}
$$

```{r}
data_pib_long <- data_pib_munic %>%
  rename(geocode      = Cód.,
         municipality = Município) %>%
  mutate(geocode = as.character(geocode)) %>%
  pivot_longer(cols      = `2002`:`2023`,
               names_to  = "year",
               values_to = "pib_thousand_brl") %>%
  mutate(
    year    = as.integer(year),
    pib_brl = as.numeric(pib_thousand_brl) * 1000
  ) %>%
  left_join(data_ipca_deflator, by = "year") %>%
  mutate(
    pib_brl_2024 = pib_brl * ipca_deflator
  ) %>%
  arrange(geocode, year)
```

I then merge the deflated GDP panel with the population panel:

```{r}
data_pib_per_capita_panel <- data_pib_long %>%
  left_join(
    data_population_panel %>% select(geocode, year, population),
    by = c("geocode", "year")
  )
```

#### 3.2.4. Handle the Mojuí dos Campos municipality split

A structural data challenge arises from the creation of **Mojuí dos Campos** (IBGE geocode `1504752`) in 2012, when it was split off from **Santarém** (geocode `1506807`), in the state of Pará. Prior to 2013, Mojuí dos Campos did not exist as an independent administrative unit and therefore has no historical GDP or population records in the IBGE series. Leaving it as `NA` for the pre-2013 period would introduce a spurious discontinuity in the panel and render the municipality unusable for the analysis.

To address this, I backcast GDP and population for Mojuí dos Campos for years prior to 2013 by applying the municipality's 2022 share of the Santarém–Mojuí combined total to Santarém's historical values. I use 2022 — the most recent census year with reliable, independently measured data for both municipalities — as the reference year for computing these shares. Santarém's pre-2013 values are simultaneously adjusted downward to subtract the imputed Mojuí contribution, ensuring that the two municipalities sum to the correct historical aggregate throughout the panel.

Since the dataset contains both nominal and deflated GDP series, I compute separate 2022 shares for each: $\hat{s}^{GDP}$ for nominal values and $\hat{s}^{GDP,2024}$ for deflated values. Population shares ($\hat{s}^{pop}$) are identical across both series. Formally, for any year $t < 2013$:

$$
GDP^{Mojuí}_{t} = GDP^{Santarém}_{t} \times \hat{s}^{GDP}, \qquad GDP^{Mojuí,2024}_{t} = GDP^{Santarém,2024}_{t} \times \hat{s}^{GDP,2024}
$$

$$
GDP^{Santarém*}_{t} = GDP^{Santarém}_{t} \times (1 - \hat{s}^{GDP}), \qquad GDP^{Santarém*,2024}_{t} = GDP^{Santarém,2024}_{t} \times (1 - \hat{s}^{GDP,2024})
$$

$$
Pop^{Mojuí}_{t} = Pop^{Santarém}_{t} \times \hat{s}^{pop}, \qquad Pop^{Santarém*}_{t} = Pop^{Santarém}_{t} \times (1 - \hat{s}^{pop})
$$

This is the same backcasting strategy applied to the Bolsa Família and PEA datasets elsewhere in this appendix (see @sec-bolsafamilia).

The operations are split across two sequential `mutate()` calls to avoid within-mutate column reference ambiguity in dplyr: the first call backcasts Mojuí's values from Santarém's original pre-split records; the second subtracts the imputed Mojuí share from Santarém's records.

```{r}
id_mojui    <- "1504752"
id_santarem <- "1506807"

shares_2022 <- data_pib_per_capita_panel %>%
  filter(year == 2022, geocode %in% c(id_mojui, id_santarem)) %>%
  summarise(
    pop_mojui      = population[geocode == id_mojui],
    pop_sant       = population[geocode == id_santarem],
    gdp_mojui      = pib_brl[geocode == id_mojui],
    gdp_sant       = pib_brl[geocode == id_santarem],
    gdp_mojui_2024 = pib_brl_2024[geocode == id_mojui],
    gdp_sant_2024  = pib_brl_2024[geocode == id_santarem]
  ) %>%
  mutate(
    pop_share_mojui      = pop_mojui / (pop_mojui + pop_sant),
    gdp_share_mojui      = gdp_mojui / (gdp_mojui + gdp_sant),
    gdp_share_mojui_2024 = gdp_mojui_2024 / (gdp_mojui_2024 + gdp_sant_2024)
  )

data_pib_adjusted <- data_pib_per_capita_panel %>%
  left_join(shares_2022, by = character()) %>%
  group_by(year) %>%
  # Step A: backcast Mojuí from Santarém's original values
  mutate(
    pib_brl = case_when(
      geocode == id_mojui & year < 2013 ~
        pib_brl[geocode == id_santarem] * gdp_share_mojui,
      TRUE ~ pib_brl
    ),
    pib_brl_2024 = case_when(
      geocode == id_mojui & year < 2013 ~
        pib_brl_2024[geocode == id_santarem] * gdp_share_mojui_2024,
      TRUE ~ pib_brl_2024
    ),
    population = case_when(
      geocode == id_mojui & year < 2013 ~
        population[geocode == id_santarem] * pop_share_mojui,
      TRUE ~ population
    )
  ) %>%
  # Step B: subtract Mojuí's imputed share from Santarém's original values
  mutate(
    pib_brl = case_when(
      geocode == id_santarem & year < 2013 ~
        pib_brl * (1 - gdp_share_mojui),
      TRUE ~ pib_brl
    ),
    pib_brl_2024 = case_when(
      geocode == id_santarem & year < 2013 ~
        pib_brl_2024 * (1 - gdp_share_mojui_2024),
      TRUE ~ pib_brl_2024
    ),
    population = case_when(
      geocode == id_santarem & year < 2013 ~
        population * (1 - pop_share_mojui),
      TRUE ~ population
    )
  ) %>%
  ungroup() %>%
  filter(year <= 2022)
```

#### 3.2.5. Construct GDP per capita and log transformation

I compute GDP per capita as the ratio of real GDP (in 2024 Reais) to municipal population, and apply a natural logarithm transformation to compress the distribution and reduce the influence of extreme values, which is standard practice for income variables in panel analyses:

$$
GDPpc^{2024}_{it} = \frac{GDP^{2024}_{it}}{Population_{it}}
$$

$$
\ln GDPpc^{2024}_{it} = \ln\left(\frac{GDP^{2024}_{it}}{Population_{it}}\right)
$$

where $GDP^{2024}_{it}$ is real municipal GDP in 2024 Reais and $Population_{it}$ is the interpolated population of municipality $i$ in year $t$.

```{r}
data_gdp_per_capita_panel <- data_pib_adjusted %>%
  mutate(
    gdp_per_capita_2024     = if_else(
      !is.na(population) & population > 0,
      pib_brl_2024 / population,
      NA_real_
    ),
    log_gdp_per_capita_2024 = log(gdp_per_capita_2024)
  ) %>%
  select(geocode, municipality, year,
         pib_brl_2024, population,
         gdp_per_capita_2024, log_gdp_per_capita_2024)
```

#### 3.2.6. Check and validate the data

I verify that the Mojuí dos Campos and Santarém series are internally consistent by confirming that the two municipalities sum to the original Santarém aggregate in all pre-2013 years, for both nominal and deflated GDP and for population. I also inspect the dataset for missing values, confirm that all geocodes are valid 7-digit IBGE identifiers, and cross-validate a random sample of observations against IBGE's online *Sistema IBGE de Recuperação Automática* (SIDRA) platform.

```{r}
# Santarém + Mojuí should recover the original combined total in all pre-2013 years
data_pib_adjusted %>%
  filter(geocode %in% c(id_mojui, id_santarem), year < 2013) %>%
  group_by(year) %>%
  summarise(
    combined_pib      = sum(pib_brl),
    combined_pib_2024 = sum(pib_brl_2024),
    combined_pop      = sum(population)
  ) %>%
  left_join(
    data_pib_per_capita_panel %>%
      filter(geocode == id_santarem, year < 2013) %>%
      select(year,
             original_pib      = pib_brl,
             original_pib_2024 = pib_brl_2024,
             original_pop      = population),
    by = "year"
  ) %>%
  mutate(
    check_pib      = round(combined_pib      - original_pib,      2),
    check_pib_2024 = round(combined_pib_2024 - original_pib_2024, 2),
    check_pop      = round(combined_pop      - original_pop,       2)
  ) %>%
  select(year, check_pib, check_pib_2024, check_pop) %>%
  print()
# All three check columns should be 0 (or negligible floating-point error)
```

### 3.3. Electoral competitiveness

#### 3.3.1. Understand the data

The electoral data used in this study comes from the **Tribunal Superior Eleitoral (TSE)**, Brazil's supreme electoral authority, which publishes official results for all federal, state, and municipal elections. The raw microdata was accessed through the **Base dos Dados** initiative [@basedosdados], a Brazilian open-data infrastructure project that standardizes, documents, and makes publicly available a wide range of administrative and governmental datasets. Base dos Dados pre-processes the TSE raw files to harmonize variable names, resolve encoding inconsistencies, and assign stable municipality identifiers compatible with the IBGE geocode standard.

I use the municipal election results for mayoral races (*cargo = "prefeito"*) from the **ordinary elections** (*eleição ordinária*) held in 2000, 2004, 2008, 2012, 2016, and 2020. The raw dataset is structured at the voting section level (*seção eleitoral*), meaning each row records the number of votes received by a given candidate in a given voting section of a given municipality. The key variables are: `ano` (election year), `id_municipio` (7-digit IBGE municipality code), `id_candidato_bd` (unique candidate identifier assigned by Base dos Dados), `sigla_partido` (party abbreviation), `turno` (round), `tipo_eleicao` (election type), and `votos` (votes cast).

#### 3.3.2. Clean and transform the data

The construction of the electoral competitiveness variables follows three sequential aggregation steps.

**Step 1 — Filter and pre-process.** I restrict the data to first-round results (`turno == 1`) of ordinary elections (`tipo_eleicao == "eleicao ordinaria"`), which drops supplementary elections (*eleições suplementares*) that are not comparable to regular electoral cycles. This choice follows the standard approach in the Brazilian electoral competition literature [@chamon2019; @arvate2013], which uses first-round vote shares to ensure comparability across all municipalities, given that only municipalities above 200,000 registered voters are subject to a second-round runoff rule. I also convert the `votos` variable from character to integer, as it is stored as a string in the raw TSE files.

**Step 2 — Aggregate to the candidate $\times$ municipality $\times$ year level.** Since the raw data is recorded at the voting-section level, I sum votes across all sections within each municipality for each candidate, identified by `id_candidato_bd`. This produces one row per candidate per municipality per election year, with the variable `candidate_votes` representing total votes received by that candidate in that municipality.

**Step 3 — Compute vote shares.** I calculate the total number of valid votes cast in each municipality-year as the sum of `candidate_votes` across all candidates, and then divide each candidate's total votes by that sum to obtain their individual vote share:

$$
s_{kit} = \frac{V_{kit}}{\sum_{j=1}^{N_{it}} V_{jit}}
$$

where $V_{kit}$ is the total votes received by candidate $k$ in municipality $i$ at election year $t$, and $N_{it}$ is the number of candidates running in that municipality-year.

```{r}
# Step 1: Filter and pre-process
mayors <- data_elections_raw %>%
  filter(
    cargo        == "prefeito",
    turno        == "1",
    tipo_eleicao == "eleicao ordinaria"
  ) %>%
  mutate(
    year  = as.integer(ano),
    votes = as.integer(votos)
  )

# Step 2: Aggregate to candidate × municipality × year
candidates <- mayors %>%
  group_by(year, id_municipio, sigla_uf, id_candidato_bd,
           numero_candidato, sigla_partido, numero_partido) %>%
  summarise(
    candidate_votes = sum(votes, na.rm = TRUE),
    .groups = "drop"
  )

# Step 3: Compute vote shares
candidates <- candidates %>%
  group_by(year, id_municipio) %>%
  mutate(
    total_votes = sum(candidate_votes),
    vote_share  = candidate_votes / total_votes
  ) %>%
  ungroup()
```

#### 3.3.3. Construct the competitiveness variables

From the candidate-level dataset, I collapse to the municipality $\times$ year level and construct five variables, described below.

##### Effective Number of Candidates (ENC)

The **Effective Number of Candidates** ($ENC$) is the standard measure of electoral fragmentation in the political economy literature [@laakso1979; @chamon2019; @arvate2013]. It is defined as the inverse of the Herfindahl-Hirschman Index (HHI) of vote-share concentration:

$$
ENC_{it} = \frac{1}{\sum_{k=1}^{N_{it}} s_{kit}^{2}}
$$

where $s_{kit}$ is the vote share of candidate $k$ in municipality $i$ at election $t$. The ENC equals 1 when a single candidate runs uncontested and increases as both the number of candidates and the evenness of vote distribution grow. It thus captures not only the raw count of competitors but also how competitively their votes are distributed, giving greater weight to races where multiple candidates hold substantial shares. An ENC of 2, for instance, indicates the electoral equivalent of two equally matched candidates, regardless of how many nominal candidates appear on the ballot.

##### Margin of Victory (MoV)

The **Margin of Victory** ($MoV$) is defined as the difference in vote shares between the first- and second-placed candidates:

$$
MoV_{it} = s_{(1)it} - s_{(2)it}
$$

where $s_{(1)it}$ and $s_{(2)it}$ are the vote shares of the winner and the runner-up in municipality $i$ at election $t$, respectively. Smaller values indicate closer races and therefore greater electoral pressure on the winner. Unlike winner vote share alone, MoV correctly classifies fragmented races where the winner holds a low absolute share but faces no single strong challenger, and closely contested two-candidate races where the winner holds a majority but barely defeated one opponent. This distinction is substantively relevant: a mayor who won with 36% of the vote in a four-candidate race may face less re-election risk than one who won with 53% against a single strong rival.

##### Competitive dummy

I construct a binary indicator equal to 1 if the election was competitive, defined as a race in which the margin of victory is below 10 percentage points:

$$
\text{Competitive}_{it} = \mathbb{1}[MoV_{it} < 0.10]
$$

This threshold is standard in the literature [@ballotpedia] and identifies elections where the outcome was genuinely uncertain. The dummy is used in robustness checks and descriptive analyses; the continuous $MoV$ is preferred in the main specifications to avoid information loss at the threshold.

##### Uncontested dummy

I flag municipalities where only one candidate ran for mayor:

$$
\text{Uncontested}_{it} = \mathbb{1}[N_{it} = 1]
$$

For uncontested races, $MoV$ is undefined (no runner-up exists) and is set to `NA`. $ENC$ is set to 1 by construction. These observations are retained in the dataset but excluded from analyses where electoral competition is the variable of interest, as they represent the structural absence of competition rather than a particular degree of it.

```{r}
data_competitiveness <- candidates %>%
  group_by(year, id_municipio, sigla_uf) %>%
  arrange(year, id_municipio, desc(vote_share)) %>%
  summarise(
    enc          = 1 / sum(vote_share^2),
    mov          = vote_share[1] - vote_share[2],
    winner_share = vote_share[1],
    n_candidates = n(),
    total_votes  = first(total_votes),
    winner_party = sigla_partido[1],
    winner_id    = id_candidato_bd[1],
    .groups      = "drop"
  ) %>%
  mutate(
    competitive  = as.integer(mov < 0.10),
    uncontested  = as.integer(n_candidates == 1),
    mov          = if_else(uncontested == 1, NA_real_, mov),
    enc          = if_else(uncontested == 1, 1, enc)
  )
```

#### 3.3.4. Check and validate the data

The final dataset contains 5,113 municipality-year observations across six election cycles (2000, 2004, 2008, 2012, 2016, and 2020). I verify the absence of duplicate municipality-year records and confirm that all `id_municipio` values match valid 7-digit IBGE geocodes. I also inspect the distribution of each variable by election year to identify implausible values and confirm consistency over time.

I cross-validate a random sample of municipality-year observations against the official TSE results published on the TSE website (<https://resultados.tse.jus.br/oficial/app/index.html>), confirming that the aggregated vote totals and winner identities are consistent with the official records.

@tbl-competitiveness-summary presents descriptive statistics for the continuous competitiveness measures, by election year.

```{r}
#| label: tbl-competitiveness-summary
#| tbl-cap: "Descriptive statistics for electoral competitiveness variables, by election year"

data_competitiveness %>%
  filter(uncontested == 0) %>%
  group_by(year) %>%
  summarise(
    N             = n(),
    `ENC (mean)`  = round(mean(enc,  na.rm = TRUE), 2),
    `ENC (sd)`    = round(sd(enc,    na.rm = TRUE), 2),
    `MoV (mean)`  = round(mean(mov,  na.rm = TRUE), 3),
    `MoV (sd)`    = round(sd(mov,    na.rm = TRUE), 3),
    `% Competitive` = round(100 * mean(competitive, na.rm = TRUE), 1),
    `% Uncontested` = round(100 * mean(uncontested), 1)
  ) %>%
  knitr::kable()
```


#### 3.3.5. Expand to a yearly panel and carry forward electoral values

The competitiveness dataset produced in the previous steps contains one observation per municipality per **election year** (2000, 2004, 2008, 2012, 2016, and 2020). Since the dissertation analysis requires a balanced municipality-year panel covering the full period from 2000 to 2020, I expand the dataset to include all intervening years and carry the electoral values forward across each four-year mayoral term.

The underlying rationale is straightforward: the competitiveness of the election that brought a mayor to office characterizes the political context under which that mayor governs for the entirety of their term. The ENC, MoV, and derived dummy variables computed for election year $t$ therefore apply to years $t+1$, $t+2$, and $t+3$, until the next election resets the values. No interpolation is performed between election cycles, as electoral competition has no meaningful continuous trajectory within a term.

Formally, for any non-election year $t'$ belonging to the term initiated by election $t$:

$$
ENC_{it'} = ENC_{it}, \quad MoV_{it'} = MoV_{it}, \qquad \forall \ t' \in \{t+1, t+2, t+3\}
$$

I first construct a complete panel shell containing all combinations of municipalities and years from 2000 to 2020, then left-join the competitiveness data onto it, and finally apply a downward fill within each municipality. An indicator variable `election_year` is added to distinguish observed election-year values from carried-forward ones.

```{r}
# Full municipality × year panel shell
panel_shell <- data_competitiveness %>%
  distinct(id_municipio, sigla_uf, geocode) %>%
  cross_join(tibble(year = 2000:2020))

# Join and carry forward
panel_competitiveness <- panel_shell %>%
  left_join(
    data_competitiveness %>%
      select(year, id_municipio, enc, mov, winner_share,
             n_candidates, total_votes, winner_party, winner_id,
             competitive, uncontested),
    by = c("year", "id_municipio")
  ) %>%
  arrange(id_municipio, year) %>%
  group_by(id_municipio) %>%
  fill(enc, mov, winner_share, n_candidates, total_votes,
       winner_party, winner_id, competitive, uncontested,
       .direction = "down") %>%
  ungroup() %>%
  mutate(
    election_year = as.integer(year %in% c(2000, 2004, 2008, 2012, 2016, 2020))
  )
```

Any remaining `NA` values after the fill reflect municipalities that are absent from the TSE records for a given election cycle — typically units created by municipal splits after 2000 that did not yet exist as independent electoral jurisdictions. These are treated as structural missings and retained as `NA` rather than imputed.
