---
title: "Audit: Bolsa Família (MDS)"
date: 2026-10-07
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Source: MDS (SAGI), MI Social query service behind the dados.gov.br
conjunto *bolsa-familia*, yearly municipal monthly files downloaded
2026-10-06 (`_data/raw/mds/bf-YYYY.txt`, 2004-2021 and 2023-2026); municipal
quota (estimativa de famílias pobres) from the SENARC LAI response of
protocol 71003.017363/2022-16
(`_data/raw/mds/Atende_SIC_PBF_2003 a Out_2021_Acumulado_Benef.Médio.xlsx`).
Pipeline script: `_scripts/02-social-bf.R`.
Interim output: `_data/interim/bf_panel_2004_2019.rds`.

```{r setup}
#| include: false
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
})
options(width = 110)
```

```{r read}
bf_dir <- "_data/raw/mds"
old_files <- file.path(bf_dir, sprintf("bf-%d.txt", c(2004:2021)))
new_files <- file.path(bf_dir, sprintf("bf-%d.txt", 2023:2026))
read_old <- function(f) {
  read_csv(f, col_types = cols(ibge = col_character(), anomes = col_integer(),
                               .default = col_double()), progress = FALSE) %>%
    mutate(source_file = basename(f))
}
bf_old <- map_dfr(old_files, read_old)
bf_new <- map_dfr(new_files, ~ read_csv(.x, col_types = cols(.default = col_character()),
                                        progress = FALSE) %>% mutate(source_file = basename(.x)))
bf_panel <- read_rds("_data/interim/bf_panel_2004_2019.rds")
munic_id <- read_xlsx("_data/raw/ibge/ibge_munic_id.xlsx", sheet = "Municípios",
                      col_types = c("text", "numeric", "text", "text"))
quota_file <- file.path(bf_dir, "Atende_SIC_PBF_2003 a Out_2021_Acumulado_Benef.Médio.xlsx")
quota_raw <- read_excel(quota_file, sheet = 1, skip = 6, col_types = "text",
                        .name_repair = "minimal")
```

## Q1. Are the yearly raw files complete and internally consistent?

What is checked: rows, municipalities, months, missingness, zeros,
negatives, positive families with zero transfers and duplicate keys per
file for 2004-2021; and the layout of the 2023-2026 files.

```{r q1-qa}
bf_old %>%
  group_by(source_file) %>%
  summarise(n = n(), municipalities = n_distinct(ibge), months = n_distinct(anomes),
            na_families = sum(is.na(qtd_familias_beneficiarias_bolsa_familia)),
            na_transfers = sum(is.na(valor_repassado_bolsa_familia)),
            zero_families = sum(qtd_familias_beneficiarias_bolsa_familia == 0, na.rm = TRUE),
            zero_transfers = sum(valor_repassado_bolsa_familia == 0, na.rm = TRUE),
            negative = sum(qtd_familias_beneficiarias_bolsa_familia < 0 |
                             valor_repassado_bolsa_familia < 0, na.rm = TRUE),
            families_pos_transfers_zero = sum(qtd_familias_beneficiarias_bolsa_familia > 0 &
                                                valor_repassado_bolsa_familia == 0, na.rm = TRUE),
            duplicate_keys = sum(duplicated(paste(ibge, anomes)))) %>%
  print(n = 20)
```

```{r q1-new}
names(read_csv(new_files[1], n_max = 0, col_types = cols(.default = col_character())))
bf_new %>% group_by(source_file) %>%
  summarise(n = n(), municipalities = n_distinct(codigo_ibge), months = n_distinct(anomes_s),
            first = min(anomes_s), last = max(anomes_s))
file.exists(file.path(bf_dir, "bf-2022.txt"))
```

**Finding.** Every file from 2004 to 2021 carries the same 5,571 six-digit
codes for twelve months, with no negative values and no duplicate keys.
Families are missing in a few dozen municipality-months per year until
2013 and in 11,152 rows of 2021 (the programme's last months before the
Auxílio Brasil substitution). In 2004-2006 the zero cells are rows where
both families and transfers are zero (930, 213 and 41 municipality-months;
in 2004, 28 municipalities for all twelve months and 55 for eight), which
is consistent with municipalities entering the programme during its
rollout rather than with reporting gaps; no row of those years has
positive families with zero transfers. That pattern appears only in 2020
(269 rows, April to December, 42 in June, the emergency-transfer months).
No file exists for 2022, and the 2023-2026 files use a new layout
(`codigo_ibge`, `anomes_s`, suffixed value columns, an average-benefit
column), with 2023 starting in March.
**Decision.** D-2026-10-05-b: panel truncated at 2019; the 2020-2026
evidence feeds task 1.2.
**Assertion.** `02-social-bf.R`, raw-state block (files present, columns,
twelve months per year, no duplicates, no negatives).

## Q2. Which month represents the year?

Candidates: February, June, July, September, November. Criteria, as in the
lost script: stability relative to the previous month, distance from the
yearly median, completeness, for beneficiary families and for the transfer
per family, 2004-2019.

```{r q2-decision}
candidate_months <- c(2, 6, 7, 9, 11)
bf_m <- bf_old %>%
  filter(anomes >= 200401, anomes <= 201912) %>%
  mutate(year = anomes %/% 100L, month = anomes %% 100L,
         families = qtd_familias_beneficiarias_bolsa_familia,
         per_family = if_else(families > 0, valor_repassado_bolsa_familia / families, NA_real_))
stability <- bf_m %>%
  group_by(ibge, year) %>% arrange(month, .by_group = TRUE) %>%
  mutate(fam_lag = lag(families), val_lag = lag(per_family), lag_month = lag(month)) %>%
  ungroup() %>%
  filter(month %in% candidate_months, lag_month == month - 1) %>%
  transmute(month, fam_instability = abs(families - fam_lag) / fam_lag,
            val_instability = abs(per_family - val_lag) / val_lag) %>%
  group_by(month) %>%
  summarise(fam_instability_p50 = median(fam_instability, na.rm = TRUE),
            fam_jump_20 = mean(fam_instability > 0.2, na.rm = TRUE),
            val_instability_p50 = median(val_instability, na.rm = TRUE),
            val_jump_20 = mean(val_instability > 0.2, na.rm = TRUE))
centrality <- bf_m %>%
  group_by(ibge, year) %>%
  mutate(fam_med = median(families, na.rm = TRUE), val_med = median(per_family, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(month %in% candidate_months) %>%
  group_by(month) %>%
  summarise(fam_distance = mean(abs(families - fam_med) / fam_med, na.rm = TRUE),
            val_distance = mean(abs(per_family - val_med) / val_med, na.rm = TRUE))
completeness <- bf_m %>%
  filter(month %in% candidate_months) %>%
  group_by(month) %>%
  summarise(na_fam = mean(is.na(families)), zero_fam = mean(families == 0, na.rm = TRUE),
            na_val = mean(is.na(per_family)))
stability %>% left_join(centrality, by = "month") %>% left_join(completeness, by = "month")
```

**Finding.** On the re-acquired files June is the most central month
(smallest mean distance to the yearly medians of families and of the
transfer per family, 1.9 and 1.4 percent) with completeness equal to the
other candidates, while February is the most stable month relative to
the previous one (lowest median change and lowest share of jumps above
20 percent); September and November return an infinite distance because
some municipality-years have a zero median. The inherited pair (June as
reference, February as alternative) is kept; the author may revisit the
ranking, since the lost script's prose credited June with the lowest
instability as well.
**Decision.** D-2026-10-05-a: `BF_REF_MONTH = 6`, `BF_ALT_MONTH = 2`.
**Assertion.** `02-social-bf.R`, output block (both months present for
every municipality-year).

## Q3. Is the implied transfer per family plausible across years?

```{r q3-per-family}
bf_m %>%
  filter(month == 6) %>%
  group_by(year) %>%
  summarise(p1 = quantile(per_family, .01, na.rm = TRUE), p50 = median(per_family, na.rm = TRUE),
            p99 = quantile(per_family, .99, na.rm = TRUE), max = max(per_family, na.rm = TRUE)) %>%
  print(n = 16)
bf_panel %>%
  filter(bf_ref_month == 6) %>%
  group_by(year) %>%
  summarise(families_millions = sum(bf_families_n, na.rm = TRUE) / 1e6,
            transfers_bn_brl_2024 = sum(bf_transfers_total_brl_2024, na.rm = TRUE) / 1e9,
            per_family_brl_2024 = transfers_bn_brl_2024 * 1e9 / (families_millions * 1e6)) %>%
  print(n = 16)
```

**Finding.** The nominal transfer per family dips in 2005-2007 and then
rises with each benefit revision; the real series (2024 BRL) shows the
coverage expansion to 2014 and the stabilisation afterwards, the known
trajectory of the programme; the tails stay within a few multiples of the
median.
**Decision.** none required.
**Assertion.** none (distributional).

## Q4. Do the municipal quota values look plausible?

What is checked: the LAI workbook's layout, the period-to-year mapping,
the quota ratio distribution in 2006-2007, the fourteen Mato Grosso cases
of task 1.4, and the zero-family rows of task 1.8.

```{r q4-workbook}
names(quota_raw)[1:9]
quota_munic <- quota_raw %>%
  select(UF, MUNICÍPIOS, IBGE, starts_with("ESTIMATIVA")) %>%
  filter(!is.na(IBGE), str_detect(IBGE, "^[0-9]{7}$")) %>%
  mutate(across(starts_with("ESTIMATIVA"), as.numeric))
nrow(quota_munic)
quota_munic %>% filter(IBGE == "1100015")
bf_panel %>% filter(geocode == "1100015", bf_ref_month == 6) %>% select(year, bf_quota) %>%
  pivot_wider(names_from = year, values_from = bf_quota)
```

```{r q4-outliers}
mt_cases <- c("Alta Floresta", "Apiacás", "Barra do Bugres", "Confresa", "Guarantã do Norte",
              "Juara", "Nova Canaã do Norte", "Nova Olímpia", "Pontes e Lacerda",
              "São José dos Quatro Marcos", "Santa Carmem", "Salto do Céu",
              "São Félix do Araguaia", "Nova Monte Verde")
ratio_0607 <- bf_panel %>%
  filter(bf_ref_month == 6, year %in% 2006:2007) %>%
  left_join(munic_id %>% select(geocode = munic_id, munic, uf), by = "geocode")
ratio_0607 %>%
  summarise(p50 = median(bf_families_quota_ratio, na.rm = TRUE),
            p99 = quantile(bf_families_quota_ratio, .99, na.rm = TRUE),
            above_3 = sum(bf_families_quota_ratio > 3, na.rm = TRUE),
            above_3_mt = sum(bf_families_quota_ratio > 3 & uf == "MT", na.rm = TRUE))
ratio_0607 %>%
  filter(uf == "MT", munic %in% mt_cases) %>%
  group_by(geocode, munic) %>%
  summarise(families = max(bf_families_n), quota_2006_08 = first(bf_quota),
            ratio = max(bf_families_quota_ratio), .groups = "drop") %>%
  arrange(desc(ratio))
ratio_0607 %>% filter(bf_families_quota_ratio > 3, !(uf == "MT" & munic %in% mt_cases)) %>%
  select(geocode, munic, uf, year, bf_families_n, bf_quota, bf_families_quota_ratio)
```

```{r q4-zero}
bf_panel %>% filter(bf_ref_month == 6, bf_families_n == 0) %>%
  left_join(munic_id %>% select(geocode = munic_id, munic, uf), by = "geocode") %>%
  count(year)
```

**Finding.** The workbook (header on row 7) carries one row per
municipality with a 7-digit code and four estimates of poor families, one
per period; the pipeline assigns them to 2004-2005, 2006-2008, 2009-2011
and 2012-2019, which is the constant-within-period pattern of the lost
`bf-quotas.xlsx`. The 2006-2007 ratio has a median near one (1.04) and a
99th percentile of 2.1. Sixty-nine municipality-years exceed 3, 61 of
them in Mato Grosso; the fourteen municipalities of task 1.4 are the
largest cases (3.1 to 12.5) and other Mato Grosso municipalities (Barra
do Garças among them) and a few in Mato Grosso do Sul and Paraná also
exceed 3. The values exceed 3 in the source itself: the LAI workbook is
the quota's origin, so the outliers are source-borne, not a tidy-step
error, and the fix (if any) needs an external quota series. Rows with
zero beneficiary families against a positive quota exist nationally in
June 2004 (98), 2005 (23) and 2006 (4).
**Decision.** D-2026-10-07-g (period mapping); task 1.4 stays open as a
source question; task 1.8 (zero rows) open.
**Assertion.** `02-social-bf.R`, raw-state and output blocks (header
names, 5,570 unique codes, every panel row of a listed municipality has a
quota).

## Q5. Do the six-digit MDS codes map to the IBGE geocodes, and how do Mojuí dos Campos and Santarém appear?

```{r q5-codes}
geocode_map <- munic_id %>% transmute(ibge = str_sub(munic_id, 1, 6), geocode = munic_id)
tibble(lookup_rows = nrow(munic_id), unique_prefixes = n_distinct(geocode_map$ibge),
       mds_codes = n_distinct(bf_old$ibge), matched = sum(unique(bf_old$ibge) %in% geocode_map$ibge))
bf_old %>%
  filter(ibge %in% c("150475", "150680"), anomes %% 100 == 6, anomes %/% 100 %in% 2011:2016) %>%
  select(ibge, anomes, qtd_familias_beneficiarias_bolsa_familia) %>%
  pivot_wider(names_from = ibge, values_from = qtd_familias_beneficiarias_bolsa_familia)
```

**Finding.** The seventh IBGE digit is a check digit, so the six-digit
prefix identifies every one of the 5,571 municipalities of the lookup, and
every MDS code matches one. Mojuí dos Campos carries its own beneficiary
counts from 2014 only; Santarém's earlier counts are not split, as
inherited.
**Decision.** D-2026-10-07-h.
**Assertion.** `02-social-bf.R`, raw-state block (unique prefixes, every
MDS code matched) and output block (no NA geocode).
