---
title: "Audit: Census anchors (PIA, PEA, informality)"
date: 2026-10-07
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Source: IBGE Censos Demográficos 2000, 2010 and 2022. Population by age
group from three recovered SIDRA workbooks (`_data/raw/ibge/data-censo{2000,
2010,2022}-pia.xlsx`: Dados do Universo 2000, Tabela 1378, Tabela 9514);
economically active population and position in occupation from the SIDRA
agregados API v3, queried 2026-10-06 (`_data/raw/ibge/sidra/*.json`:
Tabelas 616, 2953, 1572, 6580, 2031, 10261).
Pipeline script: `_scripts/02-social-census.R`.
Interim outputs: `_data/interim/census_pia_1999_2024.rds`,
`_data/interim/census_pea_informal_2000_2022.rds`.
Equivalence with the recovered panel is not checked here (the notebook never
reads `_data/reference/`); the `compare_outputs` report lives in the phase 1
script 02 commit message and in the decision entries it cites.

```{r setup}
#| include: false
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
})
options(width = 110)
```

```{r read}
source("_scripts/utils/read_sidra.R")
source("_scripts/utils/interpolate_census.R")
sidra_dir <- "_data/raw/ibge/sidra"
sidra <- function(pattern) {
  map_dfr(list.files(sidra_dir, pattern = pattern, full.names = TRUE), read_sidra_json)
}
is_geocode <- function(x) !is.na(x) & str_detect(x, "^[0-9]{7}$")

pia_2000_raw <- read_excel("_data/raw/ibge/data-censo2000-pia.xlsx", sheet = "Tabela",
                           col_types = "text")
pia_2010_raw <- read_excel("_data/raw/ibge/data-censo2010-pia.xlsx", sheet = "Tabela",
                           col_types = "text", col_names = FALSE, .name_repair = "minimal")
pia_2022_raw <- read_excel("_data/raw/ibge/data-censo2022-pia.xlsx", sheet = "Tabela",
                           col_types = "text", col_names = FALSE, .name_repair = "minimal")
census_pia <- read_rds("_data/interim/census_pia_1999_2024.rds")
census_pea_informal <- read_rds("_data/interim/census_pea_informal_2000_2022.rds")
sample_geo <- c("1100015", "1506807", "1504752", "5103254")
```

## Q1. What do the three PIA workbooks contain, and which rows form the PIA?

What is checked: the sheet layout of each vintage, the age groups each one
reports, whether the "Total" row is the population of all ages or the sum
of the groups listed, the symbols used, and the number of municipalities
per vintage.

```{r q1-layout}
map(c("_data/raw/ibge/data-censo2000-pia.xlsx", "_data/raw/ibge/data-censo2010-pia.xlsx",
      "_data/raw/ibge/data-censo2022-pia.xlsx"), excel_sheets)
names(pia_2000_raw)
pia_2010_raw[1:7, ]
pia_2022_raw[1:7, 1:8]
```

```{r q1-tidy}
pia_2000_long <- pia_2000_raw %>%
  fill(Estado, `Cód.`, `Brasil e Município`) %>%
  filter(is_geocode(`Cód.`)) %>%
  transmute(geocode = `Cód.`, age_group = `Grupo de idade`, value_chr = Populacao,
            value = sidra_value(Populacao, "-"))
pia_2010_long <- pia_2010_raw %>%
  set_names(c("geocode", "municipality", "age_group", "condition", "value_chr")) %>%
  fill(geocode, municipality) %>%
  filter(is_geocode(geocode) | geocode == "1") %>%
  mutate(value = sidra_value(value_chr, "-"))
age_header_2022 <- unlist(pia_2022_raw[6, ])
pia_2022_long <- pia_2022_raw %>%
  set_names(c("geocode", "municipality", "declaration", age_header_2022[-(1:3)])) %>%
  filter(is_geocode(geocode) | geocode == "1") %>%
  pivot_longer(-c(geocode, municipality, declaration), names_to = "age_group",
               values_to = "value_chr") %>%
  mutate(value = sidra_value(value_chr, "-"))

tibble(vintage = c(2000, 2010, 2022),
       municipalities = c(n_distinct(pia_2000_long$geocode),
                          sum(is_geocode(unique(pia_2010_long$geocode))),
                          sum(is_geocode(unique(pia_2022_long$geocode)))),
       age_groups = c(n_distinct(pia_2000_long$age_group) - 1,
                      n_distinct(pia_2010_long$age_group) - 1,
                      n_distinct(pia_2022_long$age_group) - 1),
       dash_cells = c(sum(pia_2000_long$value_chr == "-"), sum(pia_2010_long$value_chr == "-"),
                      sum(pia_2022_long$value_chr == "-")),
       other_symbols = c(sum(is.na(pia_2000_long$value)), sum(is.na(pia_2010_long$value)),
                         sum(is.na(pia_2022_long$value))))
list(`2000` = unique(pia_2000_long$age_group), `2010` = unique(pia_2010_long$age_group),
     `2022` = unique(pia_2022_long$age_group))
```

```{r q1-total}
total_vs_sum <- function(long, label) {
  long %>%
    filter(geocode %in% c("1", "1100015")) %>%
    group_by(geocode) %>%
    summarise(total_row = value[age_group == "Total"],
              sum_14_plus = sum(value[age_group != "Total"]), .groups = "drop") %>%
    mutate(vintage = label, .before = 1)
}
bind_rows(total_vs_sum(pia_2000_long, 2000), total_vs_sum(pia_2010_long, 2010),
          total_vs_sum(pia_2022_long, 2022))
```

**Finding.** Each workbook lists, per municipality, a "Total" row (or
column) followed by the age groups from 14 upward: nineteen five-year
groups in 2000 and 2022 (14; 15-19; ...; 100 and over) and thirteen in
2010 (14; 15-17; 18-19; 20-24; ...; 60-69; 70 and over). The "Total" is the
resident population of all ages (190.8 million for Brazil in 2010), not
the sum of the groups shown, so the working-age population is the sum of
the 14-and-over groups. The only non-numeric cell is "-", SIDRA's zero,
confined to the oldest groups of small municipalities. The vintages cover
5,507, 5,565 and 5,570 municipalities.
**Decision.** D-2026-10-07-a: `pia` = sum of the 14-and-over age groups.
**Assertion.** `02-social-census.R`, raw-state block (column names, age
sets, municipality counts, no NA after the zero rule).

## Q2. Which SIDRA cut gives each PEA anchor?

What is checked: the candidate API cuts for the lost `pea-YYYY.xlsx`
extracts. For 2000 and 2010: Tabela 616 (PEA, 10 and over; and the 10-14
group, so that 10+ minus 10-14 is PEA 15+), Tabela 2953 (PEA 10+, 2000
only) and Tabela 1572 (PEA 14+, 2010 only). For 2022: Tabela 6580, labour
force 14 and over.

```{r q2-cuts}
pea10 <- sidra("^sidra_616_pea10_2000_2010") %>% select(geocode, year, pea10 = value)
pea10to14 <- sidra("^sidra_616_pea10to14") %>% select(geocode, year, pea10to14 = value)
pea2953 <- sidra("^sidra_2953_pea_2000") %>% select(geocode, year, pea2953 = value)
pea14_1572 <- sidra("^sidra_1572_pea14_2010") %>% select(geocode, year, pea14_1572 = value)
lf14 <- sidra("^sidra_6580_labourforce14_2022") %>% select(geocode, year, lf14 = value)
cuts <- pea10 %>%
  full_join(pea10to14, by = c("geocode", "year")) %>%
  full_join(pea2953, by = c("geocode", "year")) %>%
  full_join(pea14_1572, by = c("geocode", "year")) %>%
  mutate(across(pea10:pea14_1572, ~ sidra_value(.x, c("-", "...")))) %>%
  mutate(pea15 = pea10 - pea10to14)
cuts %>% filter(geocode %in% sample_geo) %>% arrange(geocode, year)
cuts %>%
  group_by(year) %>%
  summarise(municipalities = n(), pea15_over_pea10 = median(pea15 / pea10, na.rm = TRUE),
            pea14_1572_over_pea15 = median(pea14_1572 / pea15, na.rm = TRUE),
            pea2953_over_pea10 = median(pea2953 / pea10, na.rm = TRUE))
lf14 %>% summarise(municipalities = n(), symbols = sum(!str_detect(lf14, "^[0-9]+$")))
```

**Finding.** The cuts differ by a few percent, so the choice matters:
PEA 15+ is 97 percent of PEA 10+ in 2000 and 98 percent in 2010, and the
14+ cut of Tabela 1572 is about one percent above PEA 15+. At the phase 1 check
(commit message), Tabela 616 PEA 15+ reproduces the recovered 2000
anchors to within one person for 460 of the 502 sample municipalities
(API cell rounding; 208 exact) and Tabela 6580 reproduces 2022 exactly,
while no cut of the 2010 tables reproduces the 2010 anchors, which sit
2-5 percent below PEA 15+ with a dispersion that rules out a constant
rescaling.
**Decision.** D-2026-10-07-b: PEA 15+ from Tabela 616 for 2000 and 2010
(the 2010 difference is registered as re-acquired-source drift under
D-2026-10-06-h, not chased), labour force 14+ from Tabela 6580 for 2022.
**Assertion.** `02-social-census.R`, raw-state block (classification
strings, years, municipality counts of the three files).

## Q3. How is informality defined from the position-in-occupation tables?

What is checked: the categories each table offers, the set counted as
formal, and the resulting share by year.

```{r q3-categories}
pos_2000 <- sidra("^sidra_2031_position_2000")
pos_2010 <- sidra("^sidra_2031_position_2010")
pos_2022 <- sidra("^sidra_10261_position_2022")
formal_2031 <- c("96167", "96168", "96170")
formal_10261 <- c("31722", "79367", "79369", "79370", "79372", "79375", "45934", "45936")
bind_rows(pos_2000 %>% mutate(table = "2031 (2000, 2010)"),
          pos_2022 %>% mutate(table = "10261 (2022)")) %>%
  distinct(table, category_id, category) %>%
  mutate(formal = category_id %in% c(formal_2031, formal_10261),
         total = category_id %in% c("0", "96165")) %>%
  print(n = 40)
```

```{r q3-shares}
informal_anchor <- function(positions, total_id, formal_ids) {
  positions %>%
    mutate(value = sidra_value(value, c("-", "..."))) %>%
    group_by(geocode, year) %>%
    summarise(occupied = sum(value[category_id == total_id]),
              formal = sum(value[category_id %in% formal_ids]), .groups = "drop") %>%
    mutate(informal = if_else(occupied > 0, 100 * (occupied - formal) / occupied, 0))
}
anchors <- bind_rows(informal_anchor(pos_2000, "0", formal_2031),
                     informal_anchor(pos_2010, "0", formal_2031),
                     informal_anchor(pos_2022, "96165", formal_10261))
anchors %>% filter(geocode %in% sample_geo) %>% arrange(geocode, year)
anchors %>%
  group_by(year) %>%
  summarise(municipalities = n(), national = 100 * (1 - sum(formal) / sum(occupied)),
            p10 = quantile(informal, .1), p50 = median(informal), p90 = quantile(informal, .9),
            zero_occupied = sum(occupied == 0))
```

**Finding.** Tabela 2031 (2000 and 2010) separates employees with a
signed card, military and statutory public servants, other employees
without a card, employers, self-employed, unpaid family workers and own
consumption; it does not split employers or the self-employed by CNPJ,
so those three groups are the formal set and the rest is informal.
Tabela 10261 (2022) does split them, and the formal set is the IBGE one:
employees with a signed card (private, domestic, public non-statutory,
state companies), military, statutory servants, employers with CNPJ and
self-employed with CNPJ. The national informal share falls from 55
percent in 2000 to 47 in 2010 and 39 in 2022; the municipal median is
higher (72, 66 and 51 percent). At the
phase 1 check the 2022 set reproduces the anchor the recovered panel
implies for every sample municipality, and the 2000 and 2010 sets
reproduce the recovered anchors to within 0.01 percentage points for
roughly 300 of 502 municipalities and within 0.05 for all but 27 (max
0.19); the lost extracts stored percentages rounded to two decimals and
their counts differ by a few persons from today's API cells.
**Decision.** D-2026-10-07-c: the category sets above; the share is
stored unrounded, on the 0-100 scale the lost extract used (task 1.9
rescales).
**Assertion.** `02-social-census.R`, raw-state block (category ids
present, years, "Total" in the other classifications).

## Q4. How are Mojuí dos Campos and Santarém handled before 2013?

What is checked: Mojuí's presence in each table, the 2022 labour-force
share used to split Santarém, and the adjusted anchors.

```{r q4-mojui}
tibble(table = c("616 PEA 2000/2010", "6580 LF 2022", "2031 2000/2010", "10261 2022",
                 "PIA workbook 2000", "PIA workbook 2010", "PIA workbook 2022"),
       mojui_present = c("1504752" %in% pea10$geocode, "1504752" %in% lf14$geocode,
                         "1504752" %in% pos_2000$geocode, "1504752" %in% pos_2022$geocode,
                         "1504752" %in% pia_2000_long$geocode, "1504752" %in% pia_2010_long$geocode,
                         "1504752" %in% pia_2022_long$geocode))
pair <- lf14 %>% filter(geocode %in% c("1504752", "1506807")) %>% mutate(lf14 = as.numeric(lf14))
share <- pair$lf14[pair$geocode == "1504752"] / sum(pair$lf14)
share
census_pea_informal %>%
  filter(geocode %in% c("1504752", "1506807"), year %in% c(2000, 2010, 2013, 2022))
```

```{r q4-plot}
#| fig-height: 3.5
census_pea_informal %>%
  filter(geocode %in% c("1504752", "1506807")) %>%
  pivot_longer(c(pea, informal)) %>%
  ggplot(aes(year, value, colour = geocode)) +
  geom_line() + geom_point(size = 1) +
  facet_wrap(~ name, scales = "free_y") +
  theme_minimal()
```

**Finding.** Mojuí dos Campos (emancipated from Santarém in 2013) appears
only in the 2022 tables and workbook. The pipeline gives it 2000 and 2010
PEA anchors equal to Santarém's times its 2022 share of the pair's labour
force (5.76 percent), reduces Santarém's anchors by the complement, and
copies Santarém's informality share. The PIA is not adjusted: Mojuí has
no PIA before 2022 and Santarém keeps its full population. The recovered
panel carries the adjusted Santarém PEA (phase 1 check), which corrects
the wording of D-2026-10-05-l for these two series.
**Decision.** D-2026-10-07-d.
**Assertion.** `02-social-census.R`, output block (every municipality of
the 2010 grid has a complete pea and informal series).

## Q5. Which municipalities are absent from the 2000 census, and what do they carry?

What is checked: "..." (not available) cells in the 2000 API tables,
municipalities missing from the 2000 workbook, and the values the pipeline
stores for them (task 1.6).

```{r q5-absent}
absent_2000 <- pea10 %>% filter(year == 2000, pea10 == "...") %>% pull(geocode)
absent_2000
setdiff(unique(pia_2010_long$geocode[is_geocode(pia_2010_long$geocode)]),
        unique(pia_2000_long$geocode))
new_2022 <- setdiff(unique(pia_2022_long$geocode), unique(pia_2010_long$geocode))
new_2022
census_pea_informal %>% filter(geocode %in% absent_2000, year %in% c(2000, 2005, 2010)) %>%
  arrange(geocode, year)
census_pia %>% filter(geocode %in% c(absent_2000[1:3], "1504752"),
                      year %in% c(1999, 2000, 2005, 2010, 2022)) %>%
  pivot_wider(names_from = geocode, values_from = pia)
```

**Finding.** Fifty-eight municipalities installed after the 2000 census
(the nine Mato Grosso ones of task 1.6 among them) read "..." in the 2000
API tables and are missing from the 2000 workbook. As inherited, their
2000 PEA and informality anchors are 0, so both series rise linearly
from zero to the 2010 value, and their PIA is back-extrapolated to 2000
from the 2010-2022 slope. The five municipalities created after 2010
have no PIA before 2022; the four other than Mojuí dos Campos have no PEA
or informality before 2022 either.
**Decision.** D-2026-10-07-e (reproduced on purpose; task 1.6 replaces
the zeros with NA and the back-extrapolation with official estimates).
**Assertion.** `02-social-census.R`, output block (no NA for the 2010
grid; NA only for post-2010 municipalities before 2022).

## Q6. Do the interpolated series stay plausible?

What is checked: negatives, extremes and the national totals of the
stored series; the 1999 and 2023-2024 PIA extrapolation.

```{r q6-summary}
census_pia %>%
  group_by(year) %>%
  summarise(n = n(), na = sum(is.na(pia)), negative = sum(pia < 0, na.rm = TRUE),
            total_millions = sum(pia, na.rm = TRUE) / 1e6, min = min(pia, na.rm = TRUE),
            max = max(pia, na.rm = TRUE)) %>%
  filter(year %in% c(1999, 2000, 2001, 2010, 2020, 2022, 2023, 2024))
census_pea_informal %>%
  group_by(year) %>%
  summarise(n = n(), na = sum(is.na(pea)), pea_negative = sum(pea < 0, na.rm = TRUE),
            pea_total_millions = sum(pea, na.rm = TRUE) / 1e6,
            informal_min = min(informal, na.rm = TRUE), informal_max = max(informal, na.rm = TRUE),
            informal_p50 = median(informal, na.rm = TRUE)) %>%
  filter(year %in% c(2000, 2005, 2010, 2015, 2020, 2022))
```

```{r q6-plot}
#| fig-height: 3.5
census_pia %>% filter(geocode %in% sample_geo) %>%
  ggplot(aes(year, pia, colour = geocode)) +
  geom_line() + geom_vline(xintercept = c(2000, 2010, 2022), linetype = 3) +
  theme_minimal()
```

**Finding.** No series turns negative. The PIA totals move smoothly across
the anchors and the 1999 and 2023-2024 values extend the adjacent decade's
slope, which is the inherited rule (the "flat" wording of D-2026-10-05-f
is wrong). The informality share stays within 0-100 everywhere; its
municipal median declines across the three censuses.
**Decision.** D-2026-10-07-f (linear extrapolation of the PIA ends).
**Assertion.** `02-social-census.R`, output block (non-negative, informal
within 0-100, row counts).
