---
title: "Audit: IBGE predominant biome by municipality"
date: 2026-10-06
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Source: IBGE, Bioma predominante por município para fins estatísticos,
file dated 2025-05-16 (recovered from the Claude Project).
Raw file: `_data/raw/ibge/ibge_bioma.csv` (semicolon-delimited, UTF-8 with BOM).
Pipeline script: `_scripts/01-env-deforestation.R` (biome join and sample filter).

```{r setup}
#| include: false
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
})
options(width = 100)
```

```{r read}
biome_raw <- read_delim("_data/raw/ibge/ibge_bioma.csv", delim = ";",
                        col_types = cols(.default = col_character()),
                        locale = locale(encoding = "UTF-8"), show_col_types = FALSE)
biome <- biome_raw %>% filter(str_detect(Geocódigo, "^[0-9]{7}$"))
biome_amz <- biome %>% filter(`Bioma predominante` == "Amazônia")
```

## Q1. What does the file contain, and which release is it?

What is checked: the rows that are not municipalities (the footer), what
they say about the source, and the number of municipality rows.

```{r q1}
tibble(lines = nrow(biome_raw), municipality_rows = nrow(biome),
       footer_rows = nrow(biome_raw) - nrow(biome))
biome_raw %>%
  filter(!str_detect(Geocódigo, "^[0-9]{7}$")) %>%
  select(Geocódigo, `Nome do município`) %>%
  mutate(`Nome do município` = str_trunc(`Nome do município`, 160))
```

**Finding.** 5,579 data lines: 5,572 geocoded rows and a seven-row footer.
The footer identifies the source as the IBGE Malha Municipal Digital 2022,
accessed December 2023, which resolves the release question left open in
the salvage register. The surviving script read an xlsx export of the same
list; the recovered file is the csv.
**Decision.** none required; provenance note updated 2026-10-06.
**Assertion.** `01-env-deforestation.R`, raw-state block
(`N_BIOME_FOOTER_ROWS`).

## Q2. Does the list contain 503 Amazônia municipalities with seven-digit codes?

```{r q2}
biome %>% count(`Bioma predominante`, sort = TRUE)
tibble(amazonia = nrow(biome_amz),
       distinct_geocodes = n_distinct(biome_amz$Geocódigo),
       seven_digits = all(nchar(biome_amz$Geocódigo) == 7),
       states = paste(sort(unique(biome_amz$`Sigla da UF`)), collapse = " "))
```

**Finding.** 503 municipalities have Amazônia as predominant biome, all
with unique seven-digit codes, across the nine Legal Amazon states.
**Decision.** D-2026-10-06-c (`BIOME_KEEP = "Amazônia"`,
`N_MUNIC_BIOME_EXPECTED = 503`); reopened by D-2026-10-05-m (task 1.1).
**Assertion.** `01-env-deforestation.R`, raw-state block.

## Q3. Which municipalities share a name across states?

```{r q3}
biome_amz %>%
  count(`Nome do município`) %>%
  filter(n > 1) %>%
  inner_join(biome_amz, by = "Nome do município") %>%
  select(`Nome do município`, Geocódigo, `Sigla da UF`) %>%
  arrange(`Nome do município`)
tibble(geocodes = n_distinct(biome_amz$Geocódigo),
       names = n_distinct(biome_amz$`Nome do município`))
```

**Finding.** Four names are shared by two municipalities each (Araguanã
MA/TO, Pau D'Arco PA/TO, Presidente Médici MA/RO, Rio Branco AC/MT): 503
geocodes, 499 names. Every join in the pipeline is on geocode.
**Decision.** none required.
**Assertion.** keys are `geocode` and `year` in every script and in
`compare_outputs.R`.

## Q4. Do municipality names agree between the IBGE list and MapBiomas?

```{r q4}
mapbiomas_names <- read_xlsx("_data/raw/mapbiomas/mapbiomas_munic.xlsx", sheet = "DEF_SECVEG") %>%
  distinct(geocode = as.character(geocode), municipality, state)
joined <- biome_amz %>% left_join(mapbiomas_names, by = c("Geocódigo" = "geocode"))
tibble(biome_munis = nrow(joined),
       missing_in_mapbiomas = sum(is.na(joined$municipality)),
       name_mismatches = sum(joined$`Nome do município` != joined$municipality, na.rm = TRUE))
joined %>% filter(`Nome do município` != municipality)
```

**Finding.** All 503 biome municipalities are in MapBiomas and every name
agrees. The one spelling difference noted in the pre-rebuild script
belonged to the lost xlsx export of the list.
**Decision.** none required.
**Assertion.** `01-env-deforestation.R`, raw-state block (`munic_names`).
