---
title: "Audit: MapBiomas deforestation and secondary vegetation"
date: 2026-10-06
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Source: MapBiomas, Deforestation and Secondary Vegetation statistics by
municipality, Collection 9 (August 2024), file dated 2025-04-24.
Raw file: `_data/raw/mapbiomas/mapbiomas_munic.xlsx`, sheet `DEF_SECVEG`.
Pipeline script: `_scripts/01-env-deforestation.R`.
Interim output: `_data/interim/mapbiomas_deforestation_amz_1987_2023.rds`.

```{r setup}
#| include: false
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
})
options(width = 100)
```

```{r read}
mapbiomas_raw <- read_xlsx("_data/raw/mapbiomas/mapbiomas_munic.xlsx", sheet = "DEF_SECVEG")
biome <- read_delim("_data/raw/ibge/ibge_bioma.csv", delim = ";",
                    col_types = cols(.default = col_character()),
                    locale = locale(encoding = "UTF-8"), show_col_types = FALSE) %>%
  filter(str_detect(Geocódigo, "^[0-9]{7}$"))
biome_amz <- biome %>% filter(`Bioma predominante` == "Amazônia")
defor <- read_rds("_data/interim/mapbiomas_deforestation_amz_1987_2023.rds")

AMZ_STATES <- c("Acre", "Amapá", "Amazonas", "Maranhão", "Mato Grosso",
                "Pará", "Rondônia", "Roraima", "Tocantins")
FOREST_TR <- c("Veg. Primária", "Veg. Secundária")
DEFOR_TR  <- c("Supressão Veg. Primária", "Supressão Veg. Secundária")
year_cols <- as.character(1985:2023)
```

## Q1. Is the raw sheet complete and keyed as expected?

What is checked: dimensions, distinct geocodes and their length, missing
values in the year columns, the transition classes present, and the size
of the nine-state subset the pipeline starts from.

```{r q1}
tibble(
  rows = nrow(mapbiomas_raw),
  cols = ncol(mapbiomas_raw),
  geocodes = n_distinct(mapbiomas_raw$geocode),
  geocode_lengths = paste(unique(nchar(as.character(mapbiomas_raw$geocode))), collapse = ","),
  na_in_year_cols = sum(is.na(mapbiomas_raw[, year_cols]))
)
mapbiomas_raw %>% count(transition_name)
nine <- mapbiomas_raw %>% filter(state %in% AMZ_STATES)
tibble(rows_nine_states = nrow(nine),
       geocodes_nine_states = n_distinct(nine$geocode),
       biome_munis_present = sum(biome_amz$Geocódigo %in% as.character(nine$geocode)))
```

**Finding.** 264,692 rows and 51 columns; 5,571 geocodes, all seven digits;
no missing value in the 39 year columns; eight transition classes. The nine
Amazon states hold 45,801 rows and 808 municipalities, and all 503
municipalities of the IBGE biome list are among them.
**Decision.** none required.
**Assertion.** `01-env-deforestation.R`, raw-state block
(`N_MUNIC_BR_EXPECTED`, `N_MUNIC_STATES_EXPECTED`).

## Q2. Are 1985 and 1986 zero for vegetation suppression?

What is checked: for the two suppression classes, the number of
municipalities with a nonzero value in each of the first four years and in
the last year of the series, nationally.

```{r q2}
mapbiomas_raw %>%
  filter(transition_name %in% DEFOR_TR) %>%
  select(transition_name, `1985`, `1986`, `1987`, `1988`, `2023`) %>%
  pivot_longer(-transition_name, names_to = "year", values_to = "area") %>%
  group_by(transition_name, year) %>%
  summarise(rows_nonzero = sum(area != 0), total_ha = sum(area), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = c(rows_nonzero, total_ha))
```

**Finding.** Both suppression classes are zero in every municipality in
1985 and 1986, the first two years of the Landsat series. 1987 is the
first year with observed primary suppression; secondary suppression starts
in 1989. 2023 carries values.
**Decision.** D-2026-10-06-a: `MAPBIOMAS_YEARS = 1987:2023`.
**Assertion.** `01-env-deforestation.R`, raw-state block
(`suppression_dropped_years`).

## Q3. Which level-1 classes compose the four transitions the pipeline sums?

What is checked: the land-cover classes (`class_level_1`) found under each
of the four transitions in the nine states, and the share of `area_forest`
that the Forest class represents for the biome municipalities in 2020.

```{r q3}
nine %>%
  filter(transition_name %in% c(FOREST_TR, DEFOR_TR)) %>%
  count(transition_name, class_level_1) %>%
  pivot_wider(names_from = transition_name, values_from = n, values_fill = 0)

nine %>%
  filter(transition_name %in% c(FOREST_TR, DEFOR_TR),
         as.character(geocode) %in% biome_amz$Geocódigo) %>%
  mutate(category = if_else(transition_name %in% FOREST_TR, "area_forest", "area_deforestation"),
         class_1 = if_else(class_level_1 == "1. Forest", "Forest", "other classes")) %>%
  group_by(category, class_1) %>%
  summarise(area_2020_ha = sum(`2020`), .groups = "drop_last") %>%
  mutate(share = area_2020_ha / sum(area_2020_ha)) %>%
  ungroup()
```

**Finding.** Primary vegetation rows span the classes Forest, Non Forest
Natural Formation, Non vegetated area, Water and Not Observed; secondary
vegetation rows also include Farming; the two suppression classes span
Forest and Non Forest Natural Formation. The surviving script sums all of
them, so `area_forest` is primary plus secondary vegetation of any class,
not the Forest class alone, and `area_deforestation` includes suppression
of non-forest natural formation. For the biome municipalities in 2020
the Forest class accounts for 95 percent of `area_forest` and 98 percent of
`area_deforestation`; the remainder is mostly non-forest natural formation.
**Decision.** D-2026-10-06-b reproduces the aggregate as inherited.
Whether the two aggregates should be restricted to class 1 (Forest) is
`open`: a definitional question for the author, outside the phase 2 task
list.
**Assertion.** none; the choice is encoded in
`MAPBIOMAS_FOREST_TRANSITIONS` and `MAPBIOMAS_DEFOR_TRANSITIONS`.

## Q4. What do the distributions of the two areas look like, and are the tails real?

What is checked: summary statistics and quantiles of `area_deforestation`
and `area_forest` over the biome municipalities, 1987-2023; the largest
municipality-years; the yearly total for the biome sample.

```{r q4}
defor %>%
  summarise(across(c(area_deforestation, area_forest),
                   list(min = min, p25 = ~ quantile(.x, 0.25), median = median,
                        mean = mean, p75 = ~ quantile(.x, 0.75), p95 = ~ quantile(.x, 0.95),
                        max = max))) %>%
  pivot_longer(everything(), names_to = c("variable", "stat"), names_sep = "_(?=[a-z0-9]+$)") %>%
  pivot_wider(names_from = stat, values_from = value)

defor %>%
  arrange(desc(area_deforestation)) %>%
  select(geocode, municipality, state_abbreviation, year, area_deforestation, area_forest) %>%
  head(10)

ggplot(defor, aes(x = area_deforestation)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(title = "area_deforestation, biome municipalities 1987-2023 (log scale)",
       x = "hectares", y = "municipality-years") +
  theme_minimal()

defor %>%
  group_by(year) %>%
  summarise(total_km2 = sum(area_deforestation) / 100) %>%
  ggplot(aes(x = year, y = total_km2)) +
  geom_col() +
  labs(title = "Deforestation, biome municipalities", x = NULL, y = "km2") +
  theme_minimal()
```

**Finding.** Both areas are strongly right-skewed, as expected from the
heterogeneity of municipal size. The largest municipality-years belong to
the known frontier municipalities (see table), and the yearly biome total
follows the known trajectory (peaks above 31,000 km2 in 2003 and 2004, a
trough near 11,000 km2 in 2010-2012, a rise afterward), so the tails
reflect the territory rather than record errors.
**Decision.** none required.
**Assertion.** none; distributional facts are not assertable.

## Q5. Do the aggregates match the MapBiomas platform for sampled municipalities?

What is checked: yearly deforestation for Acrelândia (AC) and Paragominas
(PA), the two municipalities the original run compared by hand with the
values shown on the MapBiomas platform.

```{r q5}
defor %>%
  filter(municipality %in% c("Acrelândia", "Paragominas")) %>%
  select(municipality, year, area_deforestation, area_forest) %>%
  pivot_wider(names_from = municipality, values_from = c(area_deforestation, area_forest)) %>%
  print(n = 40)
```

**Finding.** The tables reproduce the series the original run (2025)
compared against the platform and found consistent. The online comparison
is a manual step and was not repeated in the rebuild; the tables are kept
so that a reader can repeat it.
**Decision.** none required.
**Assertion.** none.

## Q6. Does Mojuí dos Campos carry data before its 2013 installation?

```{r q6}
defor %>%
  filter(geocode == "1504752") %>%
  summarise(years = n(), first_year = min(year), last_year = max(year),
            years_with_forest = sum(area_forest > 0))
```

**Finding.** Yes: MapBiomas uses the current municipal mesh for the whole
series, so Mojuí dos Campos (emancipated from Santarém in 2013) has values
for all 37 years. The pipeline keeps it at this stage; it is dropped in the
panel scripts because its social and economic series start in 2013.
**Decision.** D-2026-10-05-l (downstream).
**Assertion.** `N_MUNIC_BIOME_EXPECTED = 503` in `01-env-deforestation.R`.

## Q7. Which municipalities lack rows for some transition, and is the zero fill correct?

What is checked: the number of the four transitions present per biome
municipality in the raw sheet, and the rows of any municipality with
fewer than four.

```{r q7}
coverage <- nine %>%
  filter(transition_name %in% c(FOREST_TR, DEFOR_TR),
         as.character(geocode) %in% biome_amz$Geocódigo) %>%
  distinct(geocode, municipality, transition_name) %>%
  count(geocode, municipality, name = "transitions_present")
coverage %>% count(transitions_present)
incomplete <- coverage %>% filter(transitions_present < 4)
incomplete
nine %>%
  filter(geocode %in% incomplete$geocode, transition_name %in% c(FOREST_TR, DEFOR_TR)) %>%
  count(municipality, transition_name, class_level_1)
```

**Finding.** One municipality, Santa Cruz do Arari (PA), has no secondary
vegetation and no secondary suppression rows: its forest is primary
vegetation alone and its deforestation primary suppression alone, which
totals 16 hectares over the whole series. MapBiomas omits transitions with
no pixels,
so `values_fill = 0` in the wide reshape is the correct treatment.
**Decision.** none required.
**Assertion.** `01-env-deforestation.R`, output block (no NA in the area
columns).

## Q8. Are geocode-year keys unique and the output columns complete?

```{r q8}
tibble(
  rows = nrow(defor),
  expected = 503 * length(1987:2023),
  geocodes = n_distinct(defor$geocode),
  duplicate_keys = sum(duplicated(paste(defor$geocode, defor$year))),
  na_area_forest = sum(is.na(defor$area_forest)),
  na_area_deforestation = sum(is.na(defor$area_deforestation)),
  na_rate = sum(is.na(defor$deforestation_rate)),
  zero_forest_rows = sum(defor$area_forest == 0),
  na_z = sum(is.na(defor$deforestation_z))
)
```

**Finding.** 18,611 rows, 503 municipalities times 37 years, with unique
keys; no missing area; the rate is missing only where forest is zero (no
such row in this sample); the z-score is defined for every row.
**Decision.** none required. The rate is on 0-100 with the contemporaneous
forest area as denominator (D-2026-10-06-d; task 1.9 rescales it).
**Assertion.** `01-env-deforestation.R`, output block.
