---
title: "Audit: PPCDAm priority municipality lists"
date: 2026-10-06
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Source: author's compilation of the PPCDAm priority lists from the MMA
ordinances (Decreto 6.321/2007 and Portarias MMA 2008-2024), file created
2025-02-11, last modified 2025-12-18.
Raw files: `_data/raw/ppcdam/ppcdam_lists.xlsx`;
`_data/raw/ibge/ibge_munic_id.xlsx` (geocode lookup).
Pipeline script: `_scripts/01-env-ppcdam.R`.
Interim output: `_data/interim/ppcdam_list_2004_2024.rds`.

```{r setup}
#| include: false
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
})
options(width = 110)
```

```{r read}
path <- "_data/raw/ppcdam/ppcdam_lists.xlsx"
lista <- read_xlsx(path, sheet = "Lista")
lista_dummy <- read_xlsx(path, sheet = "Lista_dummy")
lista_cat <- suppressMessages(read_xlsx(path, sheet = "Lista_cat", .name_repair = "minimal"))
decretos <- read_xlsx(path, sheet = "Decretos")
munic_id <- read_xlsx("_data/raw/ibge/ibge_munic_id.xlsx", sheet = "Municípios",
                      col_types = c("text", "numeric", "text", "text"))
biome_amz <- read_delim("_data/raw/ibge/ibge_bioma.csv", delim = ";",
                        col_types = cols(.default = col_character()),
                        locale = locale(encoding = "UTF-8"), show_col_types = FALSE) %>%
  filter(`Bioma predominante` == "Amazônia")
ppcdam <- read_rds("_data/interim/ppcdam_list_2004_2024.rds")
year_cols <- as.character(2004:2024)
```

## Q1. What does the workbook contain, and which sheet is the indicator?

What is checked: the four sheets, whether `Lista` and `Lista_dummy`
differ, the values each coding takes, and the structure of the normative
record.

```{r q1}
excel_sheets(path)
tibble(lista_rows = nrow(lista), lista_cols = ncol(lista),
       lista_equals_dummy = isTRUE(all.equal(lista, lista_dummy)))
names(lista)
table(unlist(lista[, year_cols]), useNA = "ifany")
table(unlist(lista_cat[, -(1:2)]), useNA = "ifany")
decretos %>% select(Norm, Year, Ministry, n_prior, n_monitor, n_regress) %>%
  mutate(Norm = str_replace_all(Norm, "\r\n", " | "))
```

**Finding.** `Lista` is a 92 by 23 table (municipality, state, years
2004-2024) of 0/1 values with a single missing cell; `Lista_dummy` is
identical; `Lista_cat` codes the same cells as -1, 0 and 1 (a status
coding not used by the pipeline); `Decretos` lists the nineteen normative
acts with the municipalities each added (`prior`), moved to monitoring or
removed (`regress`). The surviving script read the first sheet, `Lista`.
**Decision.** D-2026-10-06-g: `PPCDAM_SHEET = "Lista"`.
**Assertion.** `01-env-ppcdam.R`, raw-state block (column names, values
in {0, 1, NA}).

## Q2. Can every municipality be geocoded from the IBGE lookup?

What is checked: the join of (name, state) to `ibge_munic_id.xlsx`; names
that exist in more than one state in the lookup, which is why the state is
part of the key.

```{r q2}
state_uf <- c(Acre = "AC", Amazonas = "AM", Maranhão = "MA", `Mato Grosso` = "MT",
              Pará = "PA", Rondônia = "RO", Roraima = "RR")
geocoded <- lista %>%
  mutate(uf = unname(state_uf[Estado])) %>%
  left_join(munic_id %>% select(munic, uf, geocode = munic_id), by = c("Município" = "munic", "uf"))
tibble(municipalities = nrow(geocoded),
       matched = sum(!is.na(geocoded$geocode)),
       distinct_geocodes = n_distinct(geocoded$geocode))
munic_id %>%
  filter(munic %in% lista$Município) %>%
  count(munic) %>% filter(n > 1) %>%
  inner_join(munic_id, by = "munic") %>%
  select(munic, uf, munic_id) %>% arrange(munic)
```

**Finding.** All 92 municipalities match a unique IBGE code on name plus
state. Several names exist in more than one state nationally (table), so
the state is a necessary part of the key; within the compilation no name
repeats.
**Decision.** D-2026-10-06-f: geocodes attached from the lookup through
`PPCDAM_STATE_UF`. Behavior change relative to the lost script, which read
a hand-entered Geocode column; confirmed by the author 2026-10-06.
**Assertion.** `01-env-ppcdam.R`, output block (`!anyNA(geocode)`,
92 distinct geocodes).

## Q3. Which listed municipalities fall outside the biome sample, and when were they listed?

What is checked: the 92 municipalities against the 503 of the IBGE biome
list, and the listing years of those outside it; the number of biome
municipalities ever listed within the current panel window (2008-2020),
Mojuí dos Campos excluded as in the panel.

```{r q3}
outside <- geocoded %>% filter(!geocode %in% biome_amz$Geocódigo)
outside %>%
  pivot_longer(all_of(year_cols), names_to = "year", values_to = "listed") %>%
  filter(listed == 1) %>%
  group_by(Município, Estado, geocode) %>%
  summarise(listed_years = paste(year, collapse = " "), .groups = "drop")
ppcdam %>%
  filter(geocode %in% biome_amz$Geocódigo, geocode != "1504752", year %in% 2008:2020) %>%
  group_by(geocode) %>%
  summarise(ever = any(ppcdam_list == 1, na.rm = TRUE)) %>%
  summarise(biome_munis_ever_listed_2008_2020 = sum(ever))
ppcdam %>%
  filter(geocode == "1504752", ppcdam_list == 1) %>%
  summarise(mojui_listed_years = paste(year, collapse = " "))
```

**Finding.** Four listed municipalities have a predominant biome other
than Amazônia (Grajaú MA; Comodoro, Paranatinga and Ribeirão Cascalheira
MT) and leave the panel with the biome filter; Mojuí dos Campos is in the
biome but dropped downstream. Among the remaining biome municipalities, 61
are listed at some point in 2008-2020, the number the pre-rebuild panel
carries.
**Decision.** none required now. The four outside municipalities return
with the Legal Amazon extension (D-2026-10-05-m, task 1.1).
**Assertion.** none in script 01; the 61 is a property of the joined
panel, checked at the phase 1 gate.

## Q4. Does the compilation match the ordinances year by year?

What is checked: the yearly count of listed municipalities in `Lista`
against the counts the `Decretos` sheet records for each normative act
(municipalities added as priority, moved to monitoring, removed).

```{r q4}
yearly <- tibble(year = year_cols, listed = colSums(lista[, year_cols], na.rm = TRUE))
acts <- decretos %>%
  group_by(Year) %>%
  summarise(acts = n(), added_prior = sum(n_prior, na.rm = TRUE),
            to_monitor = sum(n_monitor, na.rm = TRUE),
            removed = sum(n_regress, na.rm = TRUE), .groups = "drop") %>%
  mutate(year = as.character(Year)) %>% select(-Year)
yearly %>% left_join(acts, by = "year") %>%
  mutate(change = listed - lag(listed),
         implied = added_prior - to_monitor + removed) %>%
  print(n = 21)
```

**Finding.** The first list (2008) has 36 municipalities, as the decree
and Portaria 28 name, and 2009 adds the seven of Portaria 102. Read as
additions minus transfers to monitoring plus returns from monitoring
(`n_regress`), the act counts reproduce the yearly change in 2010, 2012,
2013, 2017, 2018 and 2020-2022. 2011 differs by one (the sheet has no
`n_regress` value for Portarias 138-139), and 2023-2024 cannot be
reconciled this way: those acts record 51 and 71 municipalities against
changes of 13 and 11, so they appear to restate the whole list rather than
its changes.
**Decision.** open. A full reconciliation requires the ordinance texts;
it belongs to the annex's treatment-variable section, not to phase 1. The
2011 gap and the 2023-2024 coding are the points to check first.
**Assertion.** `01-env-ppcdam.R`, output block
(`N_PPCDAM_LISTED_EXPECTED = 805`), which pins the compilation as read.

## Q5. Where is the missing cell, and how is it treated?

```{r q5}
lista %>% filter(if_any(all_of(year_cols), is.na)) %>%
  select(Município, Estado, `2019`, `2020`, `2021`, `2022`, `2023`, `2024`)
ppcdam %>% filter(is.na(ppcdam_list))
```

**Finding.** Grajaú (MA) has no value for 2024; every other cell is 0 or
1. The pipeline keeps the cell as NA. Grajaú is outside the biome sample
and 2024 is outside the current panel, so the value has no effect on the
reference comparison.
**Decision.** D-2026-10-06-g: kept NA, never recoded to 0.
**Assertion.** `01-env-ppcdam.R`, raw-state and output blocks
(`N_PPCDAM_NA_EXPECTED = 1`).
